---
title: "Matrix Operations"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Matrix Operations

This module extends our discussion of matrix operations. We begin with simulating least squares estimation before discussing matrix properties. We examine derivative properties before covering the projection and residual maker matrices. Understanding these operations is critical as they form the backbone of the Ordinary Least Squares (OLS) estimator, which we will derive and implement both manually and using built-in R functions.

You will find these examples useful as you derive the least squares estimator and examine the properties of the least squares estimator in later modules.

## Simulating Data

To illustrate how to use the least squares estimator in R, we can use "simulated" data. The use of simulated data allows us to specify the properties of the observations that are used in the estimation of the least squares coefficients.

For replication, we use the **set.seed** function.

By specifying **set.seed(1234)**, we will obtain the same results each time we run the code.

We first need to create the matrix of independent variables.

Assume that we want to 6,000 observations with 5 independent variables. We will add an additional column for the intercept, so our matrix will be $(6000 \times 6)$.

Our first step is to create the $(6000 \times 6)$ matrix. The matrix is filled with zeros.

We then fill the first column of the **X** matrix with 1's for the intercept term.

The next step is to fill columns 2 thorough 6 of the **X** matrix. We use a loop to fill these columns.

Each observation is filled with a random value from a standard normal distribution.

```{r}

# Clear memory
rm(list = ls())

# Set Seed to 1234
set.seed(1234)

# Create a (6000 x 6) matrix of 0s 
x <- matrix(0,6000,6)

# Fill the first column with 1s

x[,1] <- 1

# Loop over remaining columns
# Fill with values from standard normal
for (l in 2:6) {
    x[,l] <- rnorm(6000)
}

# Output the first three rows of the X matrix
kable(x[1:3,],
      digits = 3,
      col.names = c('X1','X2','X3','X4','X5','X6')) %>%
kable_styling(font_size = 12,
              position  = "center",
              full_width = FALSE)
```

## Estimating Coefficients

We can now specify the population coefficients, that is, as we are simulating population data, we can define the values of the population parameters or:

$$
\beta = \begin{bmatrix}{1 \\ 2 \\ 3 \\ 4 \\ 5 \\ 6} \end{bmatrix}
$$ Let the disturbance term, $\epsilon$ be distributed normally with 0 mean and constant variance equal to 1 or:

$$
\epsilon \sim N(0,1)
$$ We specify that $y$ is a linear equation equal to:

$$
y = X \beta + \epsilon
$$

The least squares estimator is defined as:

$$
b = (X^T X)^{-1} X^T y
$$

We estimate the least squares coefficients and produce a table of coefficients.

You should note that even though we have population data and know the population parameters, the least squares estimates are not equal to the population parameters.

The stochastic disturbance captures unobserved variation in the dependent variable, that is, we can not fully explain the variation in the dependent variable.

```{r example1}

set.seed(1234)
 
x <- matrix(0,6000,6)
      x[,1] <- 1
      for (l in 2:6) {
        x[,l] <- rnorm(6000)
      }

# Population coefficients

beta <- c(1,2,3,4,5,6)

# Disturbance term
epsilon <- rnorm(6000, 0, 1)

# y is equal to X*beta + epsilon

y <- x %*% beta + epsilon

# Solve for b =  (X^T X)^(-1) X^T y

bhat <- solve(t(x) %*% x) %*% t(x) %*% y

b_names <- c("b1", "b2", "b3", "b4", "b5", "b6")

bhat_named <- cbind(b_names, 
                    beta,
                    round(bhat,3))

kable(bhat_named,
      align     = 'c',
      col.names = c("Coefficient",
                    "Population Parameter",
                    "Estimate")) %>%
kable_styling(font_size = 12,
              position  = "center",
              full_width = FALSE)

```

## Estimating Using LM Function

We can also compare the least squares estimates using matrix algebra with those obtained using the **lm** function (which uses matrix algebra behind the scenes).

  

```{r example2, message=FALSE}

rm(list = ls())

set.seed(1234)

x <- matrix(0,10000,6)
      x[,1] <- 1
      for (l in 2:6) {
        x[,l] <- rnorm(10000)
        gc()
      }

beta <- c(7,8,9,10,11,12)

epsilon <- rnorm(10000, 0, 1)

# y is equal to X*beta + random disturbance

y <- x %*% beta + epsilon

# Solve for b =  (X^T X)^(-1) X^T y

bhat <- solve(t(x) %*% x) %*% t(x) %*% y

yhat <- x %*% bhat

e <- y - yhat

b_names <- c("b1", "b2", "b3", "b4", "b5", "b6")

data <- tibble(y = y[,1],
               x2 = x[,2],
               x3 = x[,3],
               x4 = x[,4],
               x5 = x[,5],
               x6 = x[,6])

lm_mod <- lm(y ~ x2 + x3 + x4 + x5 + x6, data = data)$coefficients

bhat_named <- cbind(b_names, 
                    beta, 
                    round(bhat,3),
                    round(lm_mod,3))
                    

kable(bhat_named,
      align     = 'c',
      col.names = c("Coefficient",
                    "Population Parameter",
                    "LS Estimates",
                    "LM Estimates")) %>%
kable_styling(font_size = 12,
              position  = "center",
              full_width = FALSE)
```
