---
title: "Hypothesis Testing"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

Module 6 focuses on the transition from parameter estimation to formal
statistical inference through the framework of hypothesis testing.
Students will examine the three fundamental approaches to
inference—Wald, fit-based, and Lagrange multiplier tests—to determine
whether empirical evidence supports specific theoretical restrictions on
population parameters. By mastering the application of $t$-tests for
individual coefficients and $F$-tests for joint hypotheses, and
implementing these procedures in R, students will gain the analytical
tools necessary to assess the statistical significance of economic
relationships and the relative performance of restricted versus
unrestricted models.

## Three Approaches

There are three general approaches to hypothesis testing:

-   **Wald** tests

-   **Fit based** tests

-   **Lagrange multiplier** tests

The hypothesis of interest is about the **population**, that is, we have
estimates and wish to examine questions about the population parameters.

If our hypothesis is correct, then the sample statistics should
approximate the description of the population parameters.

If our hypothesis is incorrect, then the sample statistics will deviate
from our description of the population parameters.

## Null Hypothesis

The **null hypothesis**, $H_0$, is the 'status quo' or typically the
assumption of 'no effect.'

The null hypothesis is the claim that we will 'operate under' unless
'strong evidence' otherwise.

In regression analysis, the null hypothesis is typically that the
regressor of interest has no impact on the dependent variable. In other
words,

$$
H_0: \beta_k = 0
$$

**We can only reject or fail to reject the null hypothesis given the
evidence at hand.**

In econometrics, we never "accept" the null hypothesis. We simply state
that the evidence is insufficient to disprove it.

## Alternative Hypothesis

The **alternative hypothesis** is the claim we are testing. Typically,
the alternative represents the presence of an effect, a difference, or a
relationship.

We do not test the alternative hypothesis. We test whether there is
sufficient evidence to reject the null hypothesis. In econometrics,
where the null hypothesis is often the true value of the population
parameter is equal to zero, we investigate whether, for a given sample,
there is sufficient evidence to reject the null of zero ('no effect').

In most cases, we have a two-sided null and alternative hypothesis, that
is, we are not testing a specific direction, merely whether we can
reject the null hypothesis of zero.

$$
H_A: \beta \ne 0
$$

## Wald Based Tests

The hypothesis states that $\boldsymbol{\beta}$ adheres to a restriction
or set of restrictions.

The restrictions, in general, can be stated at
$\boldsymbol{c(\beta) = 0}$.

Since $\boldsymbol{b}$ is a consistent estimator of
$\boldsymbol{\beta}$, $\boldsymbol{c(b) = 0}$ should be approximately 0.

The Wald test measures how close $\boldsymbol{c(b)}$ is to 0.

The Wald test is based on the **unrestricted** model.

The Wald test employs the unrestricted and restricted model estimates.

The test measures how close the estimated unrestricted model is to the
hypothesized restrictions.

## Fit Based Tests

Fit based tests are based on $R^2$ or the sum of squares.

The best possible fit is obtained using the unrestricted least squares
model.

Imposing the hypothesized restrictions degrades the fit of the least
squares model, the question is **how much**.

The fit based test compares $R^2$ in the unrestricted and restricted
models.

If the difference between the unrestricted and restricted models is
"small", the restrictions are valid.

If $R^2$ falls significantly when the restrictions are imposed, the
restrictions are likely invalid.

## Lagrange Multiplier Tests

The Lagrange Multiplier (LM) test is based on the **restricted** model.

Imposing **incorrect** restrictions yields a failed model.

If our hypothesis is that $\beta_1 = 0$, then the LM test imposes the
restriction and uses the residuals from the regression that **omits**
$x_1$.

If the restrictions are incorrect, then the omission of $x_1$ induces
omitted variable bias.

The residuals from the restricted regression will be correlated with the
other independent variables if the restriction is invalid.

The LM test statistic is based on the correlation between the restricted
regression residuals and the independent variables that are not omitted
from the regression.

## General Linear Hypotheses

We are imposing one or more linear restrictions on the linear
restriction model, that is, for the general linear model
$\boldsymbol{y = X\beta + \epsilon}$, we are imposing $\boldsymbol{J}$
linear restrictions.

Let $\boldsymbol{R}$ be a matrix where each row of $\boldsymbol{R}$
contains the coefficients of one of the restrictions.

$\boldsymbol{R}$ may have one or many rows.

We can thus write the $\boldsymbol{J}$ restrictions on
$\boldsymbol{\beta}$: $$\boldsymbol{R\ \beta\ =\ q}$$

The hypothesis implied by imposing the restrictions can be written as:
$$H_0:\ \boldsymbol{R\ \beta\ - \ q = 0}$$

$$H_A:\ \boldsymbol{R\ \beta\ - \ q \ne 0}$$

## One Hypothesis

Assume that the regression model is:

$$
\boldsymbol{y = \beta_1 + \beta_2 x_2 + \beta_3 x_3 + \dots + \beta_k x_k + \epsilon}
$$

and we hypothesize that $\beta_3 = 0$ so that:

$$
H_0: \beta_3 = 0
$$

$$H_A: \beta_3 \ne 0$$

This means that the **R** matrix has 0's for every element EXCEPT for
$\beta_3$
$$\boldsymbol{R} = [0\quad 0\quad 1\quad 0\quad \dots\quad 0];\quad \boldsymbol{q} = 0$$

So, we can restate the null:

$$
H_0: \boldsymbol{R \beta - q = 0}
$$

$$H_0: \boldsymbol{R\beta} = [0\quad 0\quad 1\quad 0\quad \dots\quad 0]\begin{bmatrix}{\beta_1 \\ \beta_2 \\ \beta_3 \\ \vdots \\ \beta_k}\end{bmatrix}= 0 = \boldsymbol{q}$$

If we multiply this out, we find:

$$
H_0: (0 \times\beta_1)+(0 \times \beta_2)+(1 \times \beta_3) + \dots + (0 \times \beta_k) = 0
$$

$$
H_0: \beta_3 = 0
$$

## Multiple Hypotheses

Assume that $\boldsymbol{y = X\beta + \epsilon}$ and we hypothesize that
$\beta_3 = \beta_4$ then
$$\boldsymbol{R} = [0\quad 0\quad 1\quad -1\quad 0\quad \dots\quad 0];\quad \boldsymbol{q} = 0$$

Assume that $\boldsymbol{y = X\beta + \epsilon}$ and we hypothesize that
$\beta_2 + \beta_3 + \beta_4 = 1$ then
$$\boldsymbol{R} = [0\quad 1\quad 1\quad 1\quad 0\quad \dots\quad 0];\quad \boldsymbol{q} = 1$$
Assume that $\boldsymbol{y = X\beta + \epsilon}$ and we hypothesize that
$\beta_2 = 0$, $\beta_3 = 0$, $\beta_k = 0$ then
$$\boldsymbol{R} = \begin{bmatrix}
      0 & 0 & 0 &\dots  & 0 \\
      0 & 1 & 0 & \dots  & 0 \\
      0 & 0 & 1 & \dots  & 0 \\
      \vdots & \vdots & \vdots & \ddots & \vdots \\
      0 & 0 & 0 & 0 & 1 
      \end{bmatrix}
      ;
      \quad
      \boldsymbol{q} 
      \begin{bmatrix}
      0 \\ 
      0 \\
      0 
      \end{bmatrix}$$

## What is R?

The $\boldsymbol{R}$ matrix must contain $K$ columns to be conformable
to $\boldsymbol{\beta}$.

The $\boldsymbol{R}$ matrix contains $J$ rows representing the $J$
restrictions.

The $\boldsymbol{R}$ matrix has full row rank, that is, $J \le K$.

The $\boldsymbol{R}$ matrix consists of linearly independent rows, which
implies that $J < K$.

If $K$ coefficients satisfy $J = K$ restrictions, then $\boldsymbol{R}$
is square and nonsingular.

If $\boldsymbol{R}$ is square and nonsingular then
$\boldsymbol{\beta = R^{-1}\ q}$.

The restriction $\boldsymbol{R\ \beta = q}$ imposes $J$ restrictions on
$K$ otherwise free parameters.

## The F-Statistic

Consider the set of $J$ linear restrictions in the null and alternative
hypotheses: $$H_0:\ \boldsymbol{R\ \beta\ - \ q = 0}$$
$$H_0:\ \boldsymbol{R\ \beta\ - \ q \ne 0}$$

Given $\boldsymbol{b}$, we want to compute the **discrepancy vector**
defined as $\boldsymbol{R\ b - q = m}$.

$\boldsymbol{m}$ is unlikely to be exactly zero, but the question is how
'close' is $\boldsymbol{m}$ to zero.

Because $\boldsymbol{b}$ is normally distributed and $\boldsymbol{m}$ in
a linear function of $\boldsymbol{b}$, $\boldsymbol{m}$ is normally
distributed.

If the null hypothesis cannot be rejected, then
$\boldsymbol{R b - q = 0}$ and $\boldsymbol{m}$ has a mean vector
$$E[\boldsymbol{m | X} ] = E[(\boldsymbol{R\ b - q)| X] = 0}$$
$$E[\boldsymbol{m | X} ] =  \boldsymbol{R}\ E[\boldsymbol{b|X}] - \boldsymbol{q = 0}$$
$$E[\boldsymbol{m | X} ] =  \boldsymbol{R\ \beta - q = 0}$$

## The F-Test

Under the null where $\boldsymbol{q = 0}$, the variance of the
discrepancy vector, \$\boldsymbol{m}\$, is equal to:
$$var[\boldsymbol{m|X}]    =  var(\boldsymbol{R\ b - q|X})$$
$$var[\boldsymbol{m|X}]    =  \boldsymbol{R\ [var(b|X)]\ R^T}$$
$$var[\boldsymbol{m|X}]    =  \boldsymbol{R}\ [\sigma^2 \boldsymbol{(X^T\ X)^{-1}}]\ \boldsymbol{R^T}$$

We can base a test of the null hypothesis **Wald Criterion**.

Conditioned on $\boldsymbol{X}$, the Wald-criterion based test is:
$$W   = \boldsymbol{m^T}\ (var[\boldsymbol{m|X}])^{-1}\ \boldsymbol{m}$$
$$W = \boldsymbol{(Rb - q)}^T\ [\boldsymbol{R}\ [\sigma^2 \boldsymbol{(X^T\ X)^{-1}}]\ \boldsymbol{R^T}]^{-1}\  \boldsymbol{(Rb - q)} \sim \chi^2$$

The Wald test using $\sigma^2$, however, is not feasible as we do not
observe $\sigma^2$.

We can employ $s^2$ instead of $\sigma^2$ and divide the result by $J$
to obtain an F-statistic: $$F    =  \frac{W}{J}\frac{\sigma^2}{s^2}$$

$$F[J,n-K|\boldsymbol{X}]    = \ \frac{\boldsymbol{(Rb - q)}^T\ [\boldsymbol{R}\ [s^2\ \boldsymbol{(X^T\ X)^{-1}}]^{-1}\ \boldsymbol{R^T}]\  \boldsymbol{(Rb - q)}}{J} \sim \chi^2$$
We employ unrestricted and restricted regressions to estimate the
F-statistic.

The restricted regression residuals are $e_r$.

The unrestricted regression residuals are $e_u$ so the F-statistic is:
$$F(J, n-k) = \frac{e_r^T e_r - e_u^T e_u}{e_u^T e_u}\frac{n-k}{J}$$

## F Test Example

```{r}

library(AER)
library(dplyr)
library(tidyverse)
data(mtcars)

# Estimate the unrestricted regression
model_u <- lm(mpg ~ cyl + hp + wt + gear + am, data = mtcars)

# Obtain the unrestricted residuals
e_u <- model_u$residuals

# Estimate the restricted regression
model_r <- lm(mpg ~ wt + gear + am, data = mtcars)

# Obtain the restricted residuals
e_r <- model_r$residuals

# Calculate the Residual Sum of Squares
e_u_ss <- t(e_u) %*% e_u
e_r_ss <- t(e_r) %*% e_r

# J = 2
# (N-k) = 32 - 6 where k = 5 + 1 (intercept)

F <- ((e_r_ss - e_u_ss)/ e_u_ss) * ((32-6)/2)
F

# Calculate using linearHypothesis function

model_u <- lm(mpg ~ cyl + hp + wt + gear + am, data = mtcars)

linearHypothesis(model_u, c("cyl = 0", "hp = 0"))

```

## Fit Based Test

Let $\boldsymbol{e_r}$ be the residuals from the restricted least
squares regression $\boldsymbol{y = X\ b_r}$. Then
$$\boldsymbol{e_r = y - X\ b_r = y - X\ b - X(b_r - b)}$$
$$\boldsymbol{e_r =e - X(b_r - b)}$$

We can now obtain $\boldsymbol{e_r^T e_r}$
$$\boldsymbol{e_r^T e_r} = [\boldsymbol{e - X(b_r - b)}]^T\ [\boldsymbol{e - X(b_r - b)}]$$
$$\boldsymbol{e_r^T e_r} = [\boldsymbol{e^T - (b_r - b)^T\ X^T}]\ [\boldsymbol{e - X(b_r - b)}]$$
$$\boldsymbol{e_r^T e_r} = \boldsymbol{e^Te - e^T\ X(b_r - b) - (b_r - b)^T\ X^Te + (b_r - b)^T\ X^T\ X(b_r - b)}$$

Since $\boldsymbol{e^T\ X = X^T\ e = 0}$
$$\boldsymbol{e_r^T e_r} = \boldsymbol{e^Te + (b_r - b)^T\ X^T\ X(b_r - b) \ge e^Te}$$

In other words, the restricted sum of squares will be at least as large
as the unrestricted sum of squares, if not larger.

## Fit Based Test

We employ unrestricted and restricted regressions to estimate the
F-statistic.

The restricted regression residuals are $e_r$.

The unrestricted regression residuals are $e_u$ so the F-statistic is:
$$F(J, n-k) = \frac{\boldsymbol{e_r^T e_r - e_u^T e_u}}{\boldsymbol{e_u^T e_u}} \frac{n-k}{J}$$

If one divides the numerator and denominator of the F-statistic by
$\Sigma(y_i - \bar{y})^2$, then
$$F(J, n-k) = \frac{R_u^2 - R_r^2}{1 - R_u^2}\frac{n-k}{J}$$

If the restrictions are invalid, then the loss of fit will be relatively
"large", and the F-statistic will be of sufficient magnitude to reject
the restrictions under the null hypothesis.

Note that the test of the significance of the regression assumes under
the null that all the coefficients for the independent variables are
equal to zero.

## Lagrange Multiplier Tests

Suppose that we explicitly impose the restrictions in the regression.
The **restricted** least squares estimator is obtained by:
$$min_{\boldsymbol{b_0}} = S(\boldsymbol{b_0}) = \boldsymbol{(y - X\ b_0)^T\ (y - X\ b_0)}$$

$$\text{subject to}\ \boldsymbol{R\ b_0 = q}$$ We can write a Lagrangian
function of this problem as:
$$L^*(\boldsymbol{b_0, \lambda}) = \boldsymbol{(y - X\ b_0)^T\ (y - X\ b_0)} + 2\ \boldsymbol{ \lambda^T (R\ b_0 - q)^2}$$
The solutions to $\boldsymbol{b_*\ \text{and}\ \lambda_*}$ will satisfy
the necessary conditions:
$$\frac{\partial L^*}{\partial \boldsymbol{b_*}} = -2 \boldsymbol{X^T\ (y - X\ b_*)} + 2 \boldsymbol{R^T \lambda = 0}$$
$$\frac{\partial L^*}{\partial \boldsymbol{\lambda_*}} = 2 \boldsymbol{R b_* - q} = 0$$

## Lagrange Multiplier Tests

Taking the necessary conditions, dividing by 2, and producing the
partitioned matrix equation yields:

$$\begin{bmatrix} \boldsymbol{X^TX} & \boldsymbol{R^T} \\ 
                  \boldsymbol{R} & \boldsymbol{0} \end{bmatrix} 
\begin{bmatrix} \boldsymbol{b_*} \\ \boldsymbol{\lambda_*} \end{bmatrix} = 
\begin{bmatrix} \boldsymbol{X^T y} \\ \boldsymbol{q} \end{bmatrix}$$

If the partitioned matrix is nonsingular, then the restricted least
squares estimator is the upper part of the solution:
$$\begin{bmatrix} \boldsymbol{b_*} \\ \boldsymbol{\lambda_*} \end{bmatrix} = \begin{bmatrix} \boldsymbol{X^TX} & \boldsymbol{R^T} \\ 
\boldsymbol{R} & \boldsymbol{0} \end{bmatrix}^{-1}\begin{bmatrix} \boldsymbol{X^T y} \\ \boldsymbol{q} \end{bmatrix} = \boldsymbol{A^{-1} d}$$

If $\boldsymbol{X^T X}$ is nonsingular, then explicit solutions can be
found for $\boldsymbol{b_*\ \text{and}\ \lambda_*}$:
$$\boldsymbol{b_* = b - (X^T\ X)^{-1}\ R^T[R\ (X^T\ X)^{-1}\ R^T]^{-1}(Rb - q)}$$
$$\boldsymbol{\lambda_* = [R\ (X^T\ X)^{-1}\ R^T]^{-1}(Rb - q)}$$

## Lagrange Multiplier Tests

We can form a test using $\boldsymbol{\lambda_* = 0}$:
$$\boldsymbol{\lambda_* = [R\ (X^T\ X)^{-1}\ R^T]^{-1}(Rb - q)} = 0 $$

A test that $\boldsymbol{\lambda_* = 0}$ should be equivalent to a test
of the null hypothesis.

If the restrictions do not hold, then $\boldsymbol{(Rb - q)} \ne 0$ and
thus $\boldsymbol{\lambda_* \ne 0}$

We can use a Wald test of the hypothesis that
$\boldsymbol{\lambda_* = 0}$:
$$W_{LM}    = \ \boldsymbol{(Rb - q)}^T\ [\boldsymbol{R}\ [\sigma^2\ \boldsymbol{(X^T\ X)^{-1}}]\ \boldsymbol{R^T}]^{-1}\  \boldsymbol{(Rb - q)}$$

The feasible version of the test is:
$$W_{LM}    = \ \boldsymbol{(Rb - q)}^T\ [\boldsymbol{R}\ [s^2\ \boldsymbol{(X^T\ X)^{-1}}]\ \boldsymbol{R^T}]^{-1}\  \boldsymbol{(Rb - q)}$$

## Lagrange Multiplier Tests

In practice, the Lagrange Multiplier test is easy to estimate.

The chi-squared test statistic with $J$ degrees of freedom can be
estimated as $n \times R^2$ in the regression
$\boldsymbol{e_* = y - Xb_*}$.

First, estimate the restricted regression to obtain the restricted
residuals $\boldsymbol{e_*}$.

Second, regress $\boldsymbol{e_*}$ on the independent variables from the
unrestricted regression $\boldsymbol{X}$.

Third, compute $W_LM = n \times R^2$ from this regression.

If $W_{LM} = n \times R^2 < \chi^2_{critical}$ then we fail to reject
the null that the restrictions are appropriate

If $W_{LM} = n \times R^2 \ge \chi^2_{critical}$ then we reject the null
that the restrictions are appropriate.

## Using R

```{r, message = FALSE}

library(broom)
library(lmtest)
library(sandwich)
library(stargazer)
library(nlme)

data("CASchools")

schools <- CASchools %>%
  mutate(STR         = students/teachers) %>%
  mutate(math_read   = (math+read)/2) %>%
  mutate(CompStu     = computer/students) %>%
  mutate(NegComp     = 100 - CompStu) %>%
  mutate(ExpenStu    = expenditure/students)

model_1 <- lm(math_read ~ english + income +
                ExpenStu, data = schools)
model_1_resids <- resid(model_1)

#Regress squared residuals on independent variables

model_1_r2 <- lm(model_1_resids^2 ~ english + 
                   income + ExpenStu, data = schools)

n_model_1_r2 <- nobs(model_1_r2)

g_model_1_r2 <- glance(model_1_r2)

k_model_1_r2 <- g_model_1_r2$df

r2_model_1_r2 <- g_model_1_r2$r.squared


#Obtain Chi-Squared Test Statistic at 5% level of significance

chi_test <- qchisq(.95, df = k_model_1_r2 - 1)

#Manually calculate Breusch-Pagan Test Statistics

#BP = Number of Observations times R-squared

model_1_man_bptest <- n_model_1_r2*r2_model_1_r2

#BP Test of Model 1

bptest(model_1)

model_1_man_bptest

```

## F-Test and T-Test

If we are testing a **single** linear restriction that the *jth*
coefficient is equal to a particular value, then $\boldsymbol{R}$ has a
single row with a one in the *jth* position and zeros elsewhere.

$\boldsymbol{R[s^2\ (X^T\ X)^-1]R^T}$ is the *jth* diagonal element of
the estimated covariance matrix.

$\boldsymbol{Rb - q} = (b_j - q)$ and the F-statistic is equal to:
$$F[1, n - K] = \frac{(b_j - q)^2}{Est.Var(b_j)}$$

Let $\hat{q}$ be the sample estimate of $q$, then we can define the
t-statistic as: $$t = \frac{\hat{q} - q}{se(\hat{q})}$$

It can be shown that the square of the t-statistic is equal to the
F-statistic or:
$$t = \frac{(\hat{q} - q)^2}{var(\hat{q}-q)} = \frac{\boldsymbol{(r^Tb - q)[r^T[s^2\ (X^TX)^{-1}]r]^{-1} (r^Tb - q)}}{1}$$

```{r}

#Estimate simple linear regression

model_1 <- lm(math_read ~ english + income +
                ExpenStu, data = schools)

#Generate Summary Results
summary(model_1)

#Conduct F-test if beta_english = 0
ftest <- linearHypothesis(model_1, c("english = 0"))

#Square Root F-test = abs(t-test)
sqrt(ftest$F[2])

```
