---
title: "Unbiasedness"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

In Module 5 of ECON 706, we examine the finite sample properties of the Ordinary Least Squares (OLS) estimator. Using numerical sampling experiments and R-based simulations, we verify that under the classical assumptions—linearity, full column rank, and strict exogeneity—the OLS estimator is unbiased, meaning its expected value centers exactly on the true population parameters.

We also analyze the critical impact of model specification on estimator performance. Through mathematical derivations and simulations, we demonstrate how omitting relevant variables introduces bias, while including irrelevant variables, though maintaining unbiasedness, unnecessarily increases the variance of our estimates. These lessons establish the foundational "Best Linear Unbiased Estimator" (BLUE) properties essential for reliable statistical inference.

## Population Estimates

We can use a sampling experiment to illustrate the sampling distribution of the least square estimates.

The $\boldsymbol{X}$ matrix consists of our standard column of 1's and a set of independent variables drawn from a random uniform distribution bounded between the values of 0 and 20.

The disturbance is drawn from a random normal such that $\epsilon \sim N(0, 100)$

We use the population parameters to construct $\boldsymbol{y}$ and then generate the least squares estimates using the population data.

```{r}

rm(list = ls())
nrows <- 1000000
ncols <- 3
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,2,4)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

bhat_pop


```

## Sampling Distributions

Now, we repeatedly draw a random sample from the sample data.

For each sample, we generate the least squares coefficients.

As our sample size increases and number of samples increases, the distribution of the least squares estimates 'centers' on the true population parameter.

The mean and median of the distribution is roughly equal to the true population parameter.

```{r}

#Set Sample and Repetitions

rm(list = ls())
nrows <- 1000000
ncols <- 3
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,2,4)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

popdata <- data.frame(y,x)

nsample <- 1000
nreps <- 10000

bhat <- matrix(ncol = ncols, nrow = nreps)

for(i in 1:nreps){
  sample <- popdata[sample(1:nrows,nsample),]
  bhat[i,] <- lm(y ~ X2 + X3, data = sample)$coefficients
   
}

hist(bhat[,2],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ 100000 ~ b[2] ~ Estimates), 
     xlab = bquote(b[2]), 
     freq = F)

```

## Unbiasedness

The sampling distribution of $\boldsymbol{b_2}$ and $\boldsymbol{b_3}$ highlight some important points.

Each least square's coefficient is, in fact, a **random** variable drawn from an underlying distribution.

With repeated sampling, the central tendency of the distribution of each parameter estimate equals the population parameter.

The parameter estimates appear to be **unbiased**, that is, on average, the parameter estimates are equal to the population parameters.

Each distribution displays variance, underlying the point that the least squares estimates are random variables.

The parameter estimates also appear to follow a normal distribution as $nreps$ and $N$ increase.

```{r}

#Set Sample and Repetitions

rm(list = ls())
nrows <- 1000000
ncols <- 3
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,2,4)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

popdata <- data.frame(y,x)

nsample <- 1000
nreps <- 10000

bhat <- matrix(ncol = ncols, nrow = nreps)

for(i in 1:nreps){
  sample <- popdata[sample(1:nrows,nsample),]
  bhat[i,] <- lm(y ~ X2 + X3, data = sample)$coefficients
   
}

hist(bhat[,3],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ 10000 ~ b[3] ~ Estimates), 
     xlab = bquote(b[3]), 
     freq = F)
```

## Finite Sample Properties

An **estimator** is a strategy, formula, or algorithm for using sample data drawn from a population to estimate the parameters in the population.

The **finite sample properties** of the least square estimator require the underlying assumptions to hold.

The finite sample properties are based on repeated sampling from the population.

The concept of **unbiasedness**, for example, depends on repeated sampling of the population.

An estimator is considered to be unbiased if the average of the estimator is equal to the population parameter.

While one estimate may be too high or too low, on average, the estimates for an unbiased estimator gravitate to the value of the population parameter.

The **asymptotic properties** of the estimator are obtained from "large" samples or the complete population.

## Unbiased Estimator

Recall, under the assumption of linearity, full column rank, and strict exogeneity that we can obtain and expand upon the least squares estimator: $$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ y}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ y = (X^T\ X)^{-1}\ X^T\ (X\ \beta + \epsilon)}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ X\ \beta + (X^T\ X)^{-1}\ X^T\ \epsilon}$$

Taking the expectation of the least squares estimator conditional on $\boldsymbol{X}$ yields $$\boldsymbol{E[b\ |\ X] = E[(X^T\ X)^{-1}\ X^T\ X\ \beta + (X^T\ X)^{-1}\ X^T\ \epsilon\ |\ X]}$$

$$\boldsymbol{E[b\ |\ X] = E[\beta] + E[(X^T\ X)^{-1}\ X^T\ \epsilon\ |\ X]}$$

$$\boldsymbol{E[b\ |\ X] = E[\beta] + (X^T\ X)^{-1}\ E[X^T\ \epsilon\ |\ X]}$$

If the assumption of strict exogeneity holds, then $\boldsymbol{E[X^T\ \epsilon\ |\ X]=0}$ and $$\boldsymbol{E[b\ |\ X] = \beta}$$

## Unbiased Estimation

Given the assumptions of the least squares estimator, we can show that the estimator is unbiased. $$\boldsymbol{E[b\ |\ X] = \beta}$$

What does this mean for a particular sample of the population?

For any sample of the population, the least squares estimator has the expectation $\boldsymbol{\beta}$. $$\boldsymbol{E[b] = E_X[E[b\ |\ x]] = \beta}$$ When we average this expectation over all the possible values of $\boldsymbol{X}$, we obtain the unconditional mean is $\boldsymbol{\beta}$.

$$\boldsymbol{E[b] = E_X[E[b\ |\ x]] = E_x[\beta] = \beta}$$

Given that the expectation of $\boldsymbol{b}$ conditional on $\boldsymbol{X}$ is equal to $\boldsymbol{\beta}$, then the least squares estimator is unbiased in **every** sample.

## Omitted Variable Bias

Let the correct model specification be:

$$\boldsymbol{y = X\ \beta + z\ \gamma + \epsilon}$$

Let $\boldsymbol{X}$ be $(n \times K)$ and $\boldsymbol{z}$ is $(n \times 1)$.

$\boldsymbol{z}$ is a relevant variable, that is, $\boldsymbol{\gamma \ne 0}$?

What happens if we omit $\boldsymbol{z}$ from the regression?

Start and expand the least squares estimator

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ y}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ (X\ \beta + z\ \gamma + \epsilon)}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ X\ \beta + (X^T\ X)^{-1}\ X^Tz\ \gamma + (X^T\ X)^{-1}\ X^T\epsilon}$$

$$\boldsymbol{b = \beta + (X^T\ X)^{-1}\ X^Tz\ \gamma + (X^T\ X)^{-1}\ X^T\epsilon}$$

## Omitted Variable Bias

Given the least squares estimator that results from the omission of the relevant variable $\boldsymbol{z}$, we can take the expectation of the estimator to show

$$\boldsymbol{b = \beta + (X^T\ X)^{-1}\ X^Tz\ \gamma + (X^T\ X)^{-1}\ X^T\epsilon}$$ $$\boldsymbol{E[b|X,z] = E[\beta|X,z] + E[(X^T\ X)^{-1}\ X^Tz\ \gamma|X,z]\ +}$$ $$\boldsymbol{E[(X^T\ X)^{-1}\ X^T\epsilon|X,z]}$$ $$\boldsymbol{E[b|X,z] = \beta + E[(X^T\ X)^{-1}\ X^Tz\ \gamma|X,z]}$$

If $\boldsymbol{X}$ and $\boldsymbol{z}$ are orthogonal, then $\boldsymbol{X^T z}= 0$.

In other words, if $\boldsymbol{z}$ is not relevant, then the least squares estimator is unbiased: $$\boldsymbol{E[b|X,z] = \beta + (X^T\ X)^{-1}\ E[X^Tz\ \gamma|X,z] = \beta}$$

But, if $\boldsymbol{X}$ and $\boldsymbol{z}$ are not orthogonal, then the least squares estimator is biased.

Let $\boldsymbol{p_{x,z}}$ equal:

$$\boldsymbol{p_{x,z} = (X^T\ X)^{-1}\ X^Tz}$$

In other words, $\boldsymbol{p_{x,z}}$ is a column vector containing the coefficients of the least squares regressions of $\boldsymbol{z}$ on $\boldsymbol{X}$.

The extent of omitted variable bias can be determined as:

$$\boldsymbol{E[b|X,z] = \beta + (X^T\ X)^{-1}\ E[X^Tz\ \gamma|X,z]}$$

$$\boldsymbol{E[b|X,z] = \beta + p_{x,z} \gamma}$$

For each coefficient in $b$, we can see that $$E[b_k|\boldsymbol{X,z} = \beta_K + \gamma \frac{Cov(z,x_k | \text{all other x's})}{Var(x_k | \text{all other x's)}}$$

While helpful, this is only helpful when there is a single included variable and a single omitted variable. As more regressors are added and omitted, the sign of the bias is no longer obvious.

## Irrelevant Variables

While omitting a relevant variable leads to omitted variable bias, what does the inclusion of an irrelevant variable do to the least squares estimator?

Assume that the true regression model is $\boldsymbol{y = X\beta + \epsilon}$ but we instead instead we estimate $\boldsymbol{y = X\ \beta + z\ \gamma + \epsilon}$.

In this case, $\boldsymbol{z}$ is an irrelevant variable, that is, $\gamma = 0$. We, however, fail to impose the restriction that $\gamma = 0$.

The least squares estimator is still unbiased.

Since $\boldsymbol{z}$ is irrelevant, $\boldsymbol{X^T z} = 0$. This implies the expectation of $\boldsymbol{b}$ is

$$\boldsymbol{E[b|X,z] = E[\beta|X,z] + E[(X^T\ X)^{-1}\ X^Tz\ \gamma|X,z]\ +}$$

$$\boldsymbol{E[(X^T\ X)^{-1}\ X^T\epsilon|X,z]}$$

$$\boldsymbol{E[b|X,z] = \beta + 0 + 0 = \beta}$$

While the least squares estimator is unbiased with the inclusion of the irrelevant variable, the variance of the estimator is adversely affected.

## Irrelevant Variables in R

```{r}

rm(list = ls())
nrows <- 100000
ncols <- 6
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,3,5,9,10,0)

y <- x %*% beta + rnorm(nrows, sd = 10)

popdata <- data.frame(y,x)

nsample <- 1000
nreps <- 1000

bhat <- matrix(ncol = ncols, nrow = nreps)

for(i in 1:nreps){
  sample <- popdata[sample(1:nrows,nsample),]
  bhat[i,] <- lm(y ~ X2 + X3 + X4 + X5 + X6, 
                 data = sample)$coefficients
   
}

mean(bhat[,2])
```

```{r, echo = FALSE, out.width = '70%', out.height= '70%'}

hist(bhat[,2],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ b[2] ~ Estimates), 
     xlab = bquote(b[2]), 
     freq = F)

hist(bhat[,6],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ b[6] ~ Estimates), 
     xlab = bquote(b[6]), 
     freq = F)


```

We can now move to the variance of the least squares estimator.
