---
title: "Variance of LS Estimator"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

In this second segment of Module 5, we shift our focus from the central
tendency of the least squares estimator to its precision and
reliability. By deriving the variance-covariance matrix of
$\boldsymbol{b}$, we establish a formal measure of the "spread" of our
estimates around the true population parameters. This derivation relies
on the assumption of homoscedasticity, allowing us to quantify how the
noise in the disturbances and the structure of our independent variables
combined to determine the stability of our coefficients.

Because the true disturbance variance, $\boldsymbol{\sigma^2}$, is
typically unknown, we introduce an unbiased estimator,
$\boldsymbol{s^2}$, derived from the residual sum of squares and the
model's degrees of freedom. This allows for the calculation of standard
errors. The module concludes with the Gauss-Markov Theorem, proving that
under the classical assumptions, the least squares estimator is the Best
Linear Unbiased Estimator (BLUE).

## Least Squares Variance

Given the standard assumptions, recall:

$$E[\boldsymbol{b|X}] = \boldsymbol{\beta}$$

The variance of a random variable, $\boldsymbol{X}$, can be expressed as
the expectation of the squared deviations from its expected value or:

$$var(\boldsymbol{X}) = E[(\boldsymbol{X} - E[\boldsymbol{X}])^2]$$

For a random vector, such as the least squares estimates,
$\boldsymbol{b}$, the concept is similar except that squaring has to be
replaced with the matrix counterpart: multiplication of a vector into
its own transpose.

The variance of the least squares estimator $\boldsymbol{b}$ can be
expressed as:
$$var(\boldsymbol{b|X}) = E[(\boldsymbol{b - E[b]})^2 | X]$$

$$var(\boldsymbol{b|X}) = E[\boldsymbol{(b - \beta)(b - \beta)^T) | X}]$$

Given that
$$\boldsymbol{b} = \boldsymbol{\beta + (X^T X)^{-1} X^T \epsilon}$$
$$\boldsymbol{b - \beta} = \boldsymbol{(X^T X)^{-1} X^T \epsilon}$$

The variance of the least squares estimator is:
$$var(\boldsymbol{b|X}) = E[\boldsymbol{(b - \beta)(b - \beta)^T) | X}]$$
$$var(\boldsymbol{b|X}) = E[\boldsymbol{[(X^T X)^{-1} X^T \epsilon]\ [(X^T X)^{-1} X^T \epsilon]^T | X}]$$
$$var(\boldsymbol{b|X}) = E[\boldsymbol{(X^T X)^{-1} X^T \epsilon \epsilon^T X (X^T X)^{-1} | X}]$$
$$var(\boldsymbol{b|X}) = \boldsymbol{(X^T X)^{-1} X^T\ E[\epsilon \epsilon^T|X]\ X (X^T X)^{-1} }$$

Recall we have assumed that
$E[\boldsymbol{\epsilon \epsilon^T | X}] = E[\boldsymbol{\epsilon \epsilon^T}] = \sigma^2 \boldsymbol{I}$.

$$var(\boldsymbol{b|X}) = \sigma^2 \boldsymbol{(X^T X)^{-1} X^TX (X^T X)^{-1}}$$
$$var(\boldsymbol{b|X}) = \sigma^2 \boldsymbol{(X^T X)^{-1}}$$

## Unknown Variance

The problem is that the variance-covariance matrix involves the
disturbance variance, $\sigma^2$, which is unknown.

Note that the residuals, $\boldsymbol{e}$ are:

$$
\boldsymbol{e = y - Xb}
$$

$$
\boldsymbol{e = y - X(X^T X)^{-1}X^Ty}
$$

$$
\boldsymbol{e = (I - X(X^T X)^{-1}X^T)y = My}
$$

$$
\boldsymbol{e = M(X \beta + \epsilon) = MX\beta + M\epsilon}
$$

Recall that $\boldsymbol{MX = 0}$ so:

$$
\boldsymbol{e = M \epsilon}
$$

We can use $\boldsymbol{e} = \boldsymbol{M \epsilon}$ to form the sum of
the squared residuals or:
$$\boldsymbol{e^T e} = \boldsymbol{(M \epsilon)^T M \epsilon}$$
$$\boldsymbol{e^T e} = \boldsymbol{\epsilon^T M^T M \epsilon}$$
$$\boldsymbol{e^T e} = \boldsymbol{\epsilon^T M \epsilon}$$

Since $\boldsymbol{\epsilon^T M \epsilon}$ is a $(1 \times 1)$ scalar,
it is equal to its **trace**.

The trace of an $(n \times n)$ square matrix $\boldsymbol{A}$ is defined
as: $$\sum_{i=1}^n a_{ii} = a_{11} + a_{22} + \dots + a_{nn}$$

In other words, the trace of a square matrix $\boldsymbol{A}$, denoted
$tr(\boldsymbol{A})$, is the sum of the elements on the main diagonal of
$\boldsymbol{A}$.

## Using the Trace

Since the trace of a scalar is just itself and
$\boldsymbol{e^T e} = \boldsymbol{\epsilon^T M \epsilon}$ is a scalar:
$$\boldsymbol{e^T e} = tr(\boldsymbol{\epsilon^T M \epsilon})$$

The expectation of $\boldsymbol{e^T e}$ conditional on $\boldsymbol{X}$
is:
$$E[\boldsymbol{e^T e|X}] = E[tr(\boldsymbol{\epsilon^T M\ \epsilon})|X]$$

Since $\boldsymbol{\epsilon^T M\ \epsilon | X}]$ is a scalar and
$\boldsymbol{M}$ is square, we can rearrange so:
$$E[\boldsymbol{e^T e|X}]  = E[tr(\boldsymbol{M\ \epsilon^T \epsilon})|X]$$

Given that $E[\boldsymbol{M |X}] = \boldsymbol{M}$, we can restate this
as:

$$E[\boldsymbol{e^T e|X}]  = tr(\boldsymbol{M}\ E[\boldsymbol{\epsilon^T \epsilon |X}])]$$

Recall the assumption of homoscedasticity means that
$E[\boldsymbol{\epsilon^T \epsilon |X}] = \sigma^2 \boldsymbol{I}$ so
that:
$$E[\boldsymbol{e^T e|X}]  = tr(\boldsymbol{M}\ \sigma^2 \boldsymbol{I})$$

Given that $\sigma^2$ is a scalar, then $tr(\sigma^2) = \sigma^2$ and we
also know that $\boldsymbol{MI} = \boldsymbol{M}$ so:
$$E[\boldsymbol{e^T e|X}]  = tr(\boldsymbol{M}\ \sigma^2 \boldsymbol{I})$$
$$E[\boldsymbol{e^T e|X}]  = \sigma^2 tr(\boldsymbol{M})$$

What is the trace of $\boldsymbol{M = I - X(X^T X)^{-1}X^T}$?

The trace of $\boldsymbol{I}$ is the sum of its diagonal elements.

Since the diagonal elements are all equal to 1 and $\boldsymbol{I}$ is
$(n \times n)$: $$tr(\boldsymbol{I}) = n$$

The trace of $\boldsymbol{X(X^T X)^{-1}X^T}$ is the trace of
$\boldsymbol{P}$.

The trace of a projection matrix is the dimension of its target space
or:
$$tr(\boldsymbol{P}) = tr(\boldsymbol{X(X^T X)^{-1}X^T}) = rank(\boldsymbol{X}) = k.$$

## S-Squared

This means that the trace of $\boldsymbol{M}$ is:

$$tr(\boldsymbol{M}) = tr(\boldsymbol{I}) - tr(\boldsymbol{X(X^T X)^{-1}X^T}) = n - k$$

Putting this all together:
$$E[\boldsymbol{e^T e|X}]  = \sigma^2 tr(\boldsymbol{M})$$

$$E[\boldsymbol{e^T e|X}]  = \sigma^2(n - K)$$ An unbiased estimator of
$\sigma^2$ is thus: $$s^2 = \frac{\boldsymbol{e^T e}}{(n-K)}$$

$s^2$ is an unbiased estimator of $\sigma^2$ conditionally and
unconditionally as:
$$E[s^2] = E_X[E[s^2|\boldsymbol{X}] = E_X[\sigma^2] = \sigma^2$$

## Variance-Covariance Matrix

Given the assumptions of linearity, full column rank, strict exogeneity,
and homoscedasticity, we know that:
$$var(\boldsymbol{b|X}) = \sigma^2 \boldsymbol{(X^T X)^{-1}}$$

We have also determined that an unbiased estimator of $\sigma^2$ is:
$$s^2 = \frac{\boldsymbol{e^T e}}{(n-K)}$$

Let $Est.Var[\boldsymbol{b|X}]$ be the sample **variance-covariance
matrix** of $\boldsymbol{b}$
$$Est.Var[\boldsymbol{b|X}] = s^2\boldsymbol{(X^T X)^{-1}}$$

The diagonal elements represent the variance of each least squares
estimator, $b_k$.

## Standard Errors

Given that the sampling variance of the least squares estimator is:

$$Est.Var[\boldsymbol{b|X}] = s^2\boldsymbol{(X^T X)^{-1}}$$

$s^2\boldsymbol{(X^T X)^{-1}}$ is square and $(k \times k)$.

If we take the square root of the diagonal elements, we obtain the
**standard errors** of the least squares estimators or:
$$se(b_k) = [[s^2\boldsymbol{(X^T X)^{-1}}]_{kk}]^{1/2}$$

The standard error of $b_k$ is akin to the standard deviation of a
randomly distributed normal variable.

For example, for $b_2$, the standard error is equal to the square root
of $s^2\boldsymbol{(X^T X)^{-1}}]_{22}$.

## Standard Errors in R

```{r}

rm(list = ls())
set.seed(1234)

nrows <- 10000
ncols <- 5
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1, 3, 5, 7, 12)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

I <- diag(nrows)
M <- (I - x^T %*% solve(t(x) %*% x) %*% t(x))
P <- x^T %*% solve(t(x) %*% x) %*% t(x)

e <- M %*% y

s2 <- as.numeric((t(e) %*% e)/(nrows - ncols))

var_cov_b <- s2 * solve(t(x) %*% x)

se_b <- sqrt(diag(var_cov_b))

var_cov_b


```

-   Our estimate for $b_2 = 2.996038$.

-   The second element on the diagonal of the variance-covariance matrix
    for $\boldsymbol{b}$ is
    $s^2(\boldsymbol{X^X}_22)^{-1} = 0.0002901079$.

-   The standard error for $b_2$ is the square root of the second
    element on the diagonal of the variance-covariance matrix for
    $\boldsymbol{b} = 0.01703256$.

## Gauss Markov Theorem

Given that $\boldsymbol{b | X} = \boldsymbol{A y}$ where
$\boldsymbol{A = (X^T X) X^T}$ is a linear function of
$\boldsymbol{\epsilon}$, then $\boldsymbol{b}$ is a linear estimator of
$\boldsymbol{\beta}$.

Because $E[\boldsymbol{A\epsilon | X}] = \boldsymbol{0}$, regardless of
the distribution of $\boldsymbol{\epsilon}$, under the other standard
assumptions, $\boldsymbol{b}$ is a linear, unbiased estimator of
$\boldsymbol{\beta}$.

Given the assumptions of Linearity, Full column rank, Strict exogeneity,
Homoscedasticity, and Stochastic or non-stochastic data or Assumptions
1-5, we can state the **Gauss-Markov Theorem**.

*Gauss-Markov Theorem*: In the linear regression model with given
regressor matrix $\boldsymbol{X}$, the least squares estimator,
$\boldsymbol{b}$ is the minimum variance unbiased estimator of
$\boldsymbol{\beta}$ and, for any vector of constants, $\boldsymbol{w}$,
the minimum variance linear unbiased estimator of
$\boldsymbol{w^T \beta}$ is $\boldsymbol{w^T b}$.

The Gauss-Markov theorem implies that under assumptions 1 - 5 that
$\boldsymbol{b}$ is the not only unbiased (on average, equal to the
population parameters) but also the most efficient (minimum variance)
estimator of $\boldsymbol{\beta}$

## Normality

While the Gauss-Markov theorem does not rely on the assumption of
normality, we often use this assumption to obtain useful results
regarding the distribution of the least square estimator.

Assume that $\boldsymbol{\epsilon}$ has a multivariate normal
distribution such that $\boldsymbol{e \sim N(0,\sigma^2)}$.

Given that $\boldsymbol{b}$ is a linear function of
$\boldsymbol{\epsilon}$ then we can use the previous results to state
$$\boldsymbol{b|X} \sim N[\boldsymbol{\beta},\sigma^2\ \boldsymbol{(X^T X)^{-1}}]$$
$$b_k|\boldsymbol{X} \sim N[\beta_k,\sigma^2\ \boldsymbol{(X^T X)_{kk}^{-1}}]$$

The exact distribution of $\boldsymbol{b}$ is conditioned on
$\boldsymbol{X}$ and the normal distribution of $\boldsymbol{b}$ in a
finite sample in a result of the specific assumption of normally
distributed disturbances.

While the normality assumption is useful for confidence intervals and
test statistics, we will be able to conduct inference about
$\boldsymbol{\beta}$ based on the sampling behavior of the statistics
independent of the normality assumption.
