---
title: "Unbiasedness"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

In Module 5 of ECON 706, we examine the finite sample properties of the Ordinary Least Squares (OLS) estimator. Using numerical sampling experiments and R-based simulations, we verify that under the classical assumptions—linearity, full column rank, and strict exogeneity—the OLS estimator is unbiased, meaning its expected value centers exactly on the true population parameters.

We also analyze the critical impact of model specification on estimator performance. Through mathematical derivations and simulations, we demonstrate how omitting relevant variables introduces bias, while including irrelevant variables, though maintaining unbiasedness, unnecessarily increases the variance of our estimates. These lessons establish the foundational "Best Linear Unbiased Estimator" (BLUE) properties essential for reliable statistical inference.

## Population Estimates

We can use a sampling experiment to illustrate the sampling distribution of the least square estimates.

The $\boldsymbol{X}$ matrix consists of our standard column of 1's and a set of independent variables drawn from a random uniform distribution bounded between the values of 0 and 20.

The disturbance is drawn from a random normal such that $\epsilon \sim N(0, 100)$

We use the population parameters to construct $\boldsymbol{y}$ and then generate the least squares estimates using the population data.

```{r}

rm(list = ls())
set.seed(1234)

nrows <- 1000000

ncols <- 3
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,2,4)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

df <- tibble(beta, bhat_pop)

kable(df,
      col.names = c('Beta', 'Beta Hat'),
      align     = 'c',
      digits    = 3) %>%
kable_classic()

```

## Sampling Distributions

In the previous code chunk, we had access to the population of observations. In reality, we must often use sample data as population data are available or costly to obtain.

**How representative are the least squares estimates when sample data are used?**

In the code chunk below, we repeatedly draw a random sample from the data.

For each sample, we generate the least squares coefficients.

We then construct a histogram of the estimates.

As our sample size increases and number of samples increases, the distribution of the least squares estimates 'centers' on the true population parameter.

The mean and median of the distribution are roughly equal to the true population parameter.

```{r}

rm(list = ls())
set.seed(1234)

nrows <- 1000000
ncols <- 3
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,2,4)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

popdata <- data.frame(y,x)

#Sample size = 1000
#Number of samples = 10000

nsample <- 1000
nreps <- 10000

bhat <- matrix(ncol = ncols, nrow = nreps)

for(i in 1:nreps){
  sample <- popdata[sample(1:nrows,nsample),]
  bhat[i,] <- lm(y ~ X2 + X3, data = sample)$coefficients
   
}

#Histogram of b2

hist(bhat[,2],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ 100000 ~ b[2] ~ Estimates), 
     xlab = bquote(b[2]), 
     freq = F)

```

## Unbiasedness

The sampling distribution of $\boldsymbol{b_2}$ suggests that the coefficients are "centered" at the true population parameter, that is, on average, the estimates are equal to the true parameter or

$$
\frac{1}{n}\sum_{i = 1}^n b_{2i} = \beta_2
$$

Each least square's coefficient is, in fact, a **random** variable drawn from an underlying distribution.

With repeated sampling, the central tendency of the distribution of each parameter estimate equals the population parameter.

The parameter estimates appear to be **unbiased**, that is, on average, the parameter estimates are equal to the population parameters.

The distribution has a variance, underlying the point that the least squares estimates are random variables.

The parameter estimates also appear to follow a normal distribution as $nreps$ and $N$ increase.

In the code chunk below, we estimate the sampling distribution for $\boldsymbol{b_3}$ with a sample size of 100 and 20,000 samples. The sampling distribution appears to be centered at $\boldsymbol{\beta_3}$. If we changed the sample size, the variance of the sampling distribution would also change.

```{r}

#Set Sample and Repetitions

rm(list = ls())
nrows <- 1000000
ncols <- 3
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,2,4)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

popdata <- data.frame(y,x)

nsample <- 100
nreps <- 20000

bhat <- matrix(ncol = ncols, nrow = nreps)

for(i in 1:nreps){
  sample <- popdata[sample(1:nrows,nsample),]
  bhat[i,] <- lm(y ~ X2 + X3, data = sample)$coefficients
   
}

hist(bhat[,3],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ 10000 ~ b[3] ~ Estimates), 
     xlab = bquote(b[3]), 
     freq = F)
```

## Finite Sample Properties

An **estimator** is a strategy, formula, or algorithm for using sample data drawn from a population to estimate the parameters in the population.

The **finite sample properties** of the least square estimator require the underlying assumptions to hold.

The finite sample properties are based on repeated sampling from the population.

The concept of **unbiasedness**, for example, depends on repeated sampling of the population.

An estimator is considered to be unbiased if the average of the estimator is equal to the population parameter.

While one estimate may be too high or too low, on average, the estimates for an unbiased estimator gravitate to the value of the population parameter.

The **asymptotic properties** of the estimator are obtained from "large" samples or the complete population.

## Unbiased Estimator

Recall, under the assumption of linearity, full column rank, and strict exogeneity that we can obtain and expand upon the least squares estimator: $$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ y}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ y = (X^T\ X)^{-1}\ X^T\ (X\ \beta + \epsilon)}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ X\ \beta + (X^T\ X)^{-1}\ X^T\ \epsilon}$$

Taking the expectation of the least squares estimator conditional on $\boldsymbol{X}$ yields $$\boldsymbol{E[b\ |\ X] = E[(X^T\ X)^{-1}\ X^T\ X\ \beta + (X^T\ X)^{-1}\ X^T\ \epsilon\ |\ X]}$$

$$\boldsymbol{E[b\ |\ X] = E[\beta] + E[(X^T\ X)^{-1}\ X^T\ \epsilon\ |\ X]}$$

$$\boldsymbol{E[b\ |\ X] = E[\beta] + (X^T\ X)^{-1}\ E[X^T\ \epsilon\ |\ X]}$$

If the assumption of strict exogeneity holds, then $\boldsymbol{E[X^T\ \epsilon\ |\ X]=0}$ and $$\boldsymbol{E[b\ |\ X] = \beta}$$

## Unbiased Estimation

Given the assumptions of the least squares estimator, we can show that the estimator is unbiased. $$\boldsymbol{E[b\ |\ X] = \beta}$$

What does this mean for a particular sample of the population?

For any sample of the population, the least squares estimator has the expectation $\boldsymbol{\beta}$. $$\boldsymbol{E[b] = E_X[E[b\ |\ x]] = \beta}$$ When we average this expectation over all the possible values of $\boldsymbol{X}$, we obtain the unconditional mean is $\boldsymbol{\beta}$.

$$\boldsymbol{E[b] = E_X[E[b\ |\ x]] = E_X[\beta] = \beta}$$

Given that the expectation of $\boldsymbol{b}$ conditional on $\boldsymbol{X}$ is equal to $\boldsymbol{\beta}$, then the least squares estimator is unbiased in **every** sample.

## Omitted Variable Bias

Let the correct model specification be:

$$\boldsymbol{y = X\ \beta + z\ \gamma + \epsilon}$$

Let $\boldsymbol{X}$ be $(n \times K)$ and $\boldsymbol{z}$ is $(n \times 1)$.

$\boldsymbol{z}$ is a relevant variable, that is, $\boldsymbol{\gamma \ne 0}$?

What happens if we omit $\boldsymbol{z}$ from the regression?

Start and expand the least squares estimator

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ y}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ (X\ \beta + z\ \gamma + \epsilon)}$$

$$\boldsymbol{b = (X^T\ X)^{-1}\ X^T\ X\ \beta + (X^T\ X)^{-1}\ X^Tz\ \gamma + (X^T\ X)^{-1}\ X^T\epsilon}$$

$$\boldsymbol{b = \beta + (X^T\ X)^{-1}\ X^Tz\ \gamma + (X^T\ X)^{-1}\ X^T\epsilon}$$

We can now take the expectation of $\boldsymbol{b}$ with respect to $\boldsymbol{X, z}$.\
\
$$\boldsymbol{E[b|X,z] = E[\beta|X,z] + E[(X^T\ X)^{-1}\ X^Tz\ \gamma|X,z] + E[(X^T\ X)^{-1}\ X^T\epsilon|X,z]}$$

$$\boldsymbol{E[b|X,z] = \beta + E[(X^T\ X)^{-1}\ X^Tz\ \gamma|X,z]}$$

If $\boldsymbol{X}$ and $\boldsymbol{z}$ are orthogonal, then $\boldsymbol{X^T z}= 0$.

In other words, if $\boldsymbol{z}$ is not relevant, then the least squares estimator is unbiased: $$\boldsymbol{E[b|X,z] = \beta + (X^T\ X)^{-1}\ E[X^Tz\ \gamma|X,z] = \beta}$$

But, if $\boldsymbol{X}$ and $\boldsymbol{z}$ are not orthogonal, then the least squares estimator is biased.

## Omitted Variables in R

In the code chunk below, we create an example of omitted variable bias.

The code chunk does the following:

-   Creates $\boldsymbol{X_1}$ as a normally distributed variable with $\mu = 50$ and $\sigma = 10$

-   Creates $\boldsymbol{X_2}$ as a variable that is highly correlated with $\boldsymbol{X_1}$

-   Defines $\beta_1 = 2$ and $\beta_2 = -1$

-   Defines $\epsilon \sim  N(0,25)$

-   Defines $y = 100 + \beta_1 x_1 + \beta_2 x_2 + \epsilon$

-   Estimates $\hat{y} = b_0 + b_1 x_1 + b_2 x_2$

-   Estimates $\hat{y} = b_0 + b_1 x_1$

-   Estimates $\hat{y} = b_0 + b_2 x_2$

-   Compares the resulting estimates

As you can see, by omitting $\boldsymbol{X_2}$ from the second regression, the estimated coefficient, $b_1$, is biased downward. When we omit $\boldsymbol{X_1}$, the estimated coefficient, $b_2$, is biased upward.

```{r}

rm(list = ls())
set.seed(1234)

library(modelsummary)

n_pop <- 100000

# Create X1 and X2 so X1 and X2 are correlated

x1 <- rnorm(n_pop, mean = 50, sd = 10)
x2 <- 0.7 * x1 + rnorm(n_pop, mean = 10, sd = 5)

beta_1 <- 2
beta_2 <- -1

epsilon <- rnorm(n_pop, mean = 0, sd = 5)

y <- 100 + (beta_1 * x1) + (beta_2 * x2) + epsilon

pop_data <- tibble(y, x1, x2)

mod_correct <- lm(y ~ x1 + x2, data = pop_data)

mod_omit_x2 <- lm(y ~ x1, data = pop_data)

mod_omit_x1 <- lm(y ~ x2, data = pop_data)

modelsummary(list(mod_correct, mod_omit_x2, mod_omit_x1))

```

## Irrelevant Variables

While omitting a relevant variable leads to omitted variable bias, what does the inclusion of an irrelevant variable do to the least squares estimator?

Let the true regression model be:

$$\boldsymbol{y = X\beta + \epsilon}$$

We add a variable, $\boldsymbol{z}$, so the regression model becomes:

$$\boldsymbol{y = X\ \beta + z\ \gamma + \epsilon}$$

Assume that $\boldsymbol{z}$ is an irrelevant variable, that is, $\gamma = 0$.

We, however, fail to impose the restriction that $\gamma = 0$.

**The least squares estimator remains unbiased.**

Since $\boldsymbol{z}$ is irrelevant, $\boldsymbol{X^T z} = 0$.

This implies the expectation of $\boldsymbol{b}$ is

$$\boldsymbol{E[b|X,z] = E[\beta|X,z] + E[(X^T\ X)^{-1}\ X^Tz\ \gamma|X,z]\ + E[(X^T\ X)^{-1}\ X^T\epsilon|X,z]}$$

$$\boldsymbol{E[b|X,z] = \beta + 0 + 0 = \beta}$$

While the least squares estimator is unbiased with the inclusion of the irrelevant variable, the variance of the estimator is adversely affected.

## Irrelevant Variables in R

```{r}

rm(list = ls())
nrows <- 100000
ncols <- 6
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1,3,5,9,10,0)

y <- x %*% beta + rnorm(nrows, sd = 10)

popdata <- data.frame(y,x)

nsample <- 1000
nreps <- 1000

bhat <- matrix(ncol = ncols, nrow = nreps)

for(i in 1:nreps){
  sample <- popdata[sample(1:nrows,nsample),]
  bhat[i,] <- lm(y ~ X2 + X3 + X4 + X5 + X6, 
                 data = sample)$coefficients
   
}

mean(bhat[,2])
```

```{r, echo = FALSE, out.width = '70%', out.height= '70%'}

hist(bhat[,2],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ b[2] ~ Estimates), 
     xlab = bquote(b[2]), 
     freq = F)

hist(bhat[,6],
 cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ b[6] ~ Estimates), 
     xlab = bquote(b[6]), 
     freq = F)


```

We can now move to the variance of the least squares estimator.
