---
title: "Robustness"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

In this section, we develop the variance estimator for the least squares
estimator. We discuss the need for a consistent estimator of the
variance of the least squares estimator. We revisit tests of the null
hypothesis of homoscedasticity.

## Robust Estimation

An estimator is **robust** if it is insensitive to violations of the
base assumptions of the model.

A **robust estimator** will retain its properties even if some of the
base assumptions that motivate the estimator are violated.

In finite samples, the unbiasedness of the least squares estimator is
not dependent on the normality assumption or even the assumption of
homoscedeastic disturbances.

Asymptotically, the least squares estimator is consistent, efficient,
and normally distributed even without the normality assumption.

**The least squares estimator is robust to the violation of the
normality assumption.**

In other words, the unbiasedness and, more importantly, the consistency
of the least squares estimator is not affected by whether the
disturbances are normally distributed (or not).

## Consistency

To establish that the least squares estimator is consistent, we first
assume $\boldsymbol{b}$ exists. In other words,
$$\frac{\boldsymbol{(X^T X)}}{n}\ \text{is full rank}$$

Second, we assume that $E[\boldsymbol{x_i}\epsilon_i]=0$ which implies
that:
$$plim_{n\to \infty} \frac{\boldsymbol{X^T \epsilon}}{n} = \boldsymbol{0}$$

Third, we assume the data are "well behaved", thus:
$$plim_{n\to \infty} \boldsymbol{\ b} = \boldsymbol{\beta} + \boldsymbol{Q^{-1}} \times \boldsymbol{0} = \boldsymbol{\beta} $$

In other words, homoscedasticity, nonautocorrelation, and normality are
not required to show the consistency of the least squares estimator and
thus $\boldsymbol{b}$ is robust to violations of these assumptions.

## Heteroscedastic Disturbances

We have previously derived, using the assumption of homoscedasticity,
the finite and asymptotic estimators of the covariance matrix of least
squares estimator or:
$$Est.Var[\boldsymbol{b}] = s^2 \boldsymbol{(X^T X)^{-1}}$$
$$Est.Asy.Var[\boldsymbol{b}] = s^2 \boldsymbol{(X^T X)^{-1}}$$
$$Asy.Var[\boldsymbol{b}] = \frac{\sigma^2}{n} \boldsymbol{Q^{-1}}$$

However, let us assume that the assumption of constant variance is
violated but the assumption of nonautocorrelation holds or:
$$Var[\epsilon_i | \boldsymbol{x_i}] = \sigma_i^2 \ne \sigma^2,\ \forall i$$
$$cov[\epsilon_i, \epsilon_j] = 0,\ \forall i \ne j$$

The question is what is the impact of the violation of the assumption
that the disturbances have constant variance on the finite sample and
asymptotic properties of the least squares estimator?

## Asymptotic Variance

Given the finite sample variance of $\boldsymbol{b}$ with
heteroscedastic disturbances is:
$$var(\boldsymbol{b|X}) = \boldsymbol{(X^T X)^{-1} \left( X^T\ \Omega\ X \right) (X^T X)^{-1}}$$

And recalling that
$$plim_{n\to \infty}\ \bigg(\frac{\boldsymbol{X^T X}}{n} \bigg)^{-1} = \boldsymbol{Q^{-1}}$$

Then what is the probability limit of the finite sample variance?
$$plim_{n\to \infty}\ var(\boldsymbol{b|X}) = plim_{n\to \infty} \left[\boldsymbol{(X^T X)^{-1} \left( X^T\ \Omega\ X \right) (X^T X)^{-1}} \right]$$
$$plim_{n\to \infty}\ var(\boldsymbol{b|X}) = plim\ \boldsymbol{(X^T X)^{-1}}\ plim \boldsymbol{(X^T\ \Omega\ X)}\ plim\ \boldsymbol{(X^T X)^{-1}}$$
$$plim_{n\to \infty}\ var(\boldsymbol{b|X}) = n \, \boldsymbol{Q^{-1}}\ plim \boldsymbol{(X^T\ \Omega\ X)}\ n \, \boldsymbol{Q^{-1}}$$

The question is can we find a consistent estimator for
$\boldsymbol{(X^T\ \Omega\ X)}$?

## White's Estimator

While we never observe the true disturbances $\boldsymbol{\epsilon}$, we
do observe the residuals $\boldsymbol{e}$.

White (1980) noted that an individual residual $e_i$ will not provide a
reasonable estimate of the corresponding disturbance $\epsilon_i$.

White (1980) argued, however, that that $\boldsymbol{(X^T e e^T X)}$ is
a **consistent** (but not unbiased) estimator of
$\boldsymbol{(X^T E[\epsilon \epsilon^T] X)}$.

Our challenge is that we want to estimate
$n^{-1} \Sigma\ \epsilon_i^2\ \boldsymbol{x_i x_i^T}$ but we don't
observe $\boldsymbol{\epsilon}$.

We can estimate $\epsilon_i$ by
$e_{i} = y_i - \hat{y_i} = y_i - \boldsymbol{x_i^T\ b}$.

We can now define the estimator of $\boldsymbol{\Omega}$ as:
$$\boldsymbol{\hat{\Omega}} = \boldsymbol{e e^T}$$

White's heteroscedasticity robust estimator of
$n^{-1} \Sigma\ \epsilon_i^2\ \boldsymbol{x_i x_i^T}$ is:
$$\boldsymbol{W_{het}} = \frac{1}{n} \Sigma\ e_i^2\ \boldsymbol{x_i x_i^T}$$

It can be shown that:
$$plim_{n\to \infty}\ e_i = plim_{n\to \infty}\ (y - \boldsymbol{x_i^T b}) = y - \boldsymbol{x_i^T \beta} = \epsilon$$
$$plim_{n\to \infty} e_i^2 = \epsilon_i^2$$

This implies that:
$$plim_{n\to \infty}\boldsymbol{W_{het}} = plim_{n\to \infty}\frac{1}{n} \Sigma\ e_i^2\ \boldsymbol{x_i x_i^T}$$
$$plim_{n\to \infty}\boldsymbol{W_{het}} = plim_{n\to \infty}\frac{1}{n} \Sigma\ \epsilon_i^2\ \boldsymbol{x_i x_i^T}$$
$$plim_{n\to \infty}\boldsymbol{W_{het}} = plim_{n\to \infty}\frac{1}{n} \Sigma\ \sigma_i^2\ \boldsymbol{x_i x_i^T}$$

**We can use the observed residuals to estimate the behavior of the
unknown disturbances.**

Recall that $\boldsymbol{X^T X = \Sigma x_i x_i^T}$ so that:
$$\boldsymbol{X^T \Omega X} = \begin{bmatrix} x_1 & x_2 & \dots & x_n \end{bmatrix}
\begin{bmatrix}\sigma_1^2 & 0 & \dots & 0 \\ 0 & \sigma_2^2 & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\
0 & 0 & \dots & \sigma_n^2 \end{bmatrix}\begin{bmatrix} x_1^T \\ x_2^T \\ \dots \\ x_n^T \end{bmatrix} = \Sigma\ \sigma_i^2\ \boldsymbol{x_i x_i^T}$$

Utilizing the least squares residuals, we obtain:
$$\boldsymbol{X^T \hat{\Omega} X} = \begin{bmatrix} x_1 & x_2 & \dots & x_n \end{bmatrix}
\begin{bmatrix}e_1^2 & 0 & \dots & 0 \\ 0 & e_2^2 & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\
0 & 0 & \dots & e_n^2 \end{bmatrix}\begin{bmatrix} x_1^T \\ x_2^T \\ \dots \\ x_n^T \end{bmatrix} = \Sigma\ e_i^2\ \boldsymbol{x_i x_i^T}$$

## White's Estimator

White's heteroscedasticity robust estimator of the variance-covariance
matrix for $\boldsymbol{b}$ is:
$$HC0.var(\boldsymbol{b|X}) = \boldsymbol{(X^T X)^{-1} \left( X^T\ \hat{\Omega}\ X \right) (X^T X)^{-1}}$$

If the regression errors are independent but have distinct variances,
$\sigma_i^2$, then:
$$\Omega = diag(\sigma_1^2, \sigma_2^2,...,\sigma_n^2)$$

And we can estimate $\Omega$ with the squared residuals or:
$$\hat{\Omega} = diag(e_1^2, e_2^2, ..., e_n^2)$$

White's variance-covariance matrix is now:

$$HC.var(\boldsymbol{b|X}) = \boldsymbol{(X^T X)^{-1} \frac{1}{n} \sum_{i=1}^n e_i^2\ \boldsymbol{x_i x_i^T}  (X^T X)^{-1}}$$
$$HC0.var(\boldsymbol{b|X}) = \boldsymbol{(X^T X)^{-1} \left( X^T\ diag[e_i^2] \ X \right) (X^T X)^{-1}}$$

## Heteroscedasticity

White (1980) shows that
$\frac{1}{n} \Sigma e_i^2\ \boldsymbol{x_i x_i^2}$ is a consistent but
biased estimator of
$\frac{1}{n} \Sigma\ \sigma^2 \boldsymbol{x_i x_i^T}$.

White is not the only method of generating heteroscedasticity consistent
standard errors.

It is now standard to employ a finite sample correction to White's
estimator to the bias.

A simple bias correcting heteroscedasticity consistent estimator is
simply:
$$HC1.var(\boldsymbol{b|X}) = \frac{n}{n-k} \boldsymbol{(X^T X)^{-1} \left( X^T\ diag[e_i^2] \ X \right) (X^T X)^{-1}}$$

It should be clear that as $n$ becomes large, the bias correction
approaches 1 for this estimator.

These estimators are asymptotically consistent.

## Homoscedastic Example

```{r, message = FALSE, warnings = FALSE}

rm(list = ls())
set.seed(1234)

library(broom)
library(lmtest)
library(sandwich)
library(stargazer)
library(nlme)

nrows <- 10000
ncols <- 5
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1, 3, 5, 7, 12)

y <- x %*% beta + rnorm(nrows, sd = 10)

bhat_pop <- solve(t(x) %*% x) %*% t(x) %*% y

I <- diag(nrows)
M <- (I - x^T %*% solve(t(x) %*% x) %*% t(x))
P <- x^T %*% solve(t(x) %*% x) %*% t(x)

e <- M %*% y

s2 <- as.numeric((t(e) %*% e)/(nrows - ncols))
var_cov_b <- s2 * solve(t(x) %*% x)
se_b <- sqrt(diag(var_cov_b))

var_cov_b

cbind(bhat_pop, matrix(se_b))

summary(lm(y ~ x[,2] + x[,3] + x[,4] + x[,5]))

```

## Heteroscedastic Example

```{r, message = FALSE, warnings = FALSE}

rm(list = ls())
set.seed(1234)

library(broom)
library(lmtest)
library(sandwich)
library(stargazer)
library(nlme)

nrows <- 10000
ncols <- 5
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- runif(nrows, min = 0, max = 20)
        gc()
      }

beta <- c(1, 3, 5, 7, 12)

epsilon_het <- rnorm(nrows, sd = x[,2])

y <- x %*% beta + rnorm(nrows, sd = 10)

y_het <- x %*% beta + epsilon_het

bhat_het <- solve(t(x) %*% x) %*% t(x) %*% y_het

I <- diag(nrows)
M <- (I - x^T %*% solve(t(x) %*% x) %*% t(x))
P <- x^T %*% solve(t(x) %*% x) %*% t(x)

e <- M %*% y

e_het <- M %*% y_het
e_het2 <- e_het^2

xex <- 0

#We need to calculate X^T ehet^2 X but it is large (NxN)
for(i in 1:nrow(x)) {
xex <- xex + e_het2[i]*x[i,]%*%t(x[i,])
}

vcov_h0 <- solve(t(x) %*% x) %*% xex %*% solve(t(x) %*% x)
se_h0 <- sqrt(diag(vcov_h0))

vcov_h1 <- nrows/(nrows - ncols) * 
           solve(t(x) %*% x) %*% xex %*% solve(t(x) %*% x)

se_h1 <- sqrt(diag(vcov_h1))


#Obtaining the variance-covariance matrix
#HC1 providers the White standard errors

model_1 <- lm(y_het ~ x[,2] + x[,3] + x[,4] + x[,5])
vcov_h0 <- vcovHC(model_1, type = "HC0")
vcov_h1 <- vcovHC(model_1, type = "HC1")

model_1_h0 <- coeftest(model_1, vcov = vcov_h0)

model_1_robust.se_h0 <- sqrt(diag(vcov_h0))
model_1_robust.se_h1 <- sqrt(diag(vcov_h1))

cbind(se_h0, model_1_robust.se_h0)

cbind(se_h1, model_1_robust.se_h1)

```

## White's Test

White (1980) proposed a test for heteroscedasticity that follows the
logic of the Bruesch-Pagan test.

White proposed to add the squares and cross-products of all the
independent variables to the regression of the squared residuals on the
independent variables.
$$e^2_i = \hat{z_0} + \hat{\delta_1} z_{i1} + ... +\ \hat{\delta_k} z_{ik} + .. +\ \hat{\lambda_2} z_{i2}^2 + ...+ \hat{\lambda_k} z_{ik}^2 + \hat{\tau}z_{i1}z_{i2} + \hat{\tau}z_{i1}z_{i3} + ... + \hat{\upsilon_i}$$

We should note that
$\hat{y} = \hat{\delta_1} z_{i1} + \hat{\delta_2} z_{i2} + ... + \hat{\delta_k} z_{ik}$.

The White test becomes
$$e^2_i = \hat{\delta_0} + \delta_1 \hat{y} + \delta_2 \hat{y}^2 + \hat{\upsilon_i}$$

Under the null (homoscedasticity), the sample size $N$ times $R^2$ is
distributed $\chi^2$ with $K-1$ degrees of freedom or
$\chi^2 = N \times R^2 \sim \chi^2_{K-1}$.

## Example

```{r, message = FALSE, warnings = FALSE}

rm(list = ls())

library(AER)
data("CASchools")

library(broom)
library(lmtest)
library(sandwich)
library(stargazer)
library(nlme)

data("CASchools")

schools <- CASchools %>%
  mutate(STR         = students/teachers) %>%
  mutate(math_read   = (math+read)/2) %>%
  mutate(CompStu     = computer/students) %>%
  mutate(NegComp     = 100 - CompStu) %>%
  mutate(ExpenStu    = expenditure/students)

model_1 <- lm(math_read ~ english + income +
                ExpenStu, data = schools)
model_1_resids <- resid(model_1)

#Regress squared residuals on independent variables

model_1_r2 <- lm(model_1_resids^2 ~ english + 
                   income + ExpenStu, data = schools)

n_model_1_r2 <- nobs(model_1_r2)

g_model_1_r2 <- glance(model_1_r2)

k_model_1_r2 <- g_model_1_r2$df

r2_model_1_r2 <- g_model_1_r2$r.squared

#Obtain Chi-Squared Test Statistic at 5% level of significance

chi_test <- qchisq(.95, df = k_model_1_r2 - 1)

#Manually calculate Breusch-Pagan Test Statistics

#BP = Number of Observations times R-squared

model_1_man_bptest <- n_model_1_r2*r2_model_1_r2

#BP Test of Model 1

bptest(model_1)

model_1_man_bptest

#Obtaining the variance-covariance matrix
#HC0 providers the White standard errors

vcov_1 <- vcovHC(model_1, type = "HC0")

model_1_robust <- coeftest(model_1, vcov = vcov_1)

model_1_robust.se <- sqrt(diag(vcov_1))


#All One Step

model_2 <- lm(math_read ~ english + income + ExpenStu, data = schools)

model_2_bptest <- bptest(model_2)

model_2_robust <- coeftest(model_2, vcov =  vcovHC(model_2, type = "HC0"))

model_2_robust.se <- sqrt(diag(vcovHC(model_2, type = "HC0")))

#White Test

model_2_white <- lm(math_read ~ english + income + ExpenStu  
                  + I(english^2) + I(income^2) + I(ExpenStu^2) 
                  + I(english*income) + I(english*ExpenStu) + I(income*ExpenStu), data = schools)

model_2_white_res <- resid(model_2_white)

model_2_white_var <- lm(model_2_white_res^2 ~ english + income + ExpenStu  
                  + I(english^2) + I(income^2) + I(ExpenStu^2) 
                  + I(english*income) + I(english*ExpenStu) + I(income*ExpenStu), data = schools)

bptest(model_2_white)
```
