---
title: "Partitioned Regression"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

Understanding partitioned regression is an element of discussing the
properties of the least squares regression model. Omitting relevant
variables or variables that are correlated with other independent
variables influences the least squares parameter estimates. The
coefficients from the multiple regression model are partial regression
coefficients.

## Orthogonal Estimates

If the matrix of independent variables can be partitioned into two
**orthogonal** matrices, then the partitioned least squares estimates
are:

$$\begin{bmatrix} \boldsymbol{b_1} \\ \boldsymbol{b_2} \end{bmatrix} = \begin{bmatrix} \boldsymbol{X_1^T X_1} & \boldsymbol{0} \\ \boldsymbol{0} & \boldsymbol{X_2^T X_2} \end{bmatrix}^{-1}  \begin{bmatrix} \boldsymbol{X_1^T y} \\ \boldsymbol{X_2^T y} \end{bmatrix}$$

$$\begin{bmatrix} \boldsymbol{b_1} \\ \boldsymbol{b_2} \end{bmatrix} = \begin{bmatrix} (\boldsymbol{X_1^T X_1})^{-1} & 0 \\ 0 & (\boldsymbol{X_2^T X_2})^{-1} \end{bmatrix} \begin{bmatrix} \boldsymbol{X_1^T y} \\ \boldsymbol{X_2^T y} \end{bmatrix}$$
$$\begin{bmatrix} \boldsymbol{b_1} \\ \boldsymbol{b_2} \end{bmatrix} = \begin{bmatrix} \boldsymbol{(X_1^T X_1)^{-1}X_1^T y} \\ \boldsymbol{(X_2^T X_2)^{-1}X_2^T y} \end{bmatrix}$$

Since the $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ are orthogonal then
$(\boldsymbol{X^T X})$ matrix is block diagonal.

Since $(\boldsymbol{X^T X})$ matrix is block diagonal, the least squares
estimates of the coefficients for the first set of independent variables
from the regression of $\boldsymbol{y}$ on $\boldsymbol{X_1}$ are the
same as the least squares estimates obtained from the regression of
$\boldsymbol{y}$ on $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$.

**This, however, is not the usual case**.

## Non-Orthogonal Partitions

It is more likely that the partition of the matrix of independent
variables will create two **non-orthogonal** matrices.

If $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ are non-orthogonal, then
the Orthogonal Partitioned Regression Theorem no longer applies. In
other words, we cannot regress $\boldsymbol{y}$ separately on
$\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ and obtain the same
coefficients that we would obtain by regressing $\boldsymbol{y}$ on
$\boldsymbol{X}$.

Start with the partitioned normal equations:

$$\begin{matrix} (1) \\ (2) \end{matrix} \begin{bmatrix} \boldsymbol{X_1^T X_1} & \boldsymbol{X_1^T X_2} \\ \boldsymbol{X_2^T X_1} & \boldsymbol{X_2^T X_2} \end{bmatrix} \begin{bmatrix} \boldsymbol{b_1} \\ \boldsymbol{b_2} \end{bmatrix} = \begin{bmatrix} \boldsymbol{X_1^T y} \\ \boldsymbol{X_2^T y} \end{bmatrix}$$

From the partitioned normal equations, we can solve (1) for
$\boldsymbol{b_1}$.

$$(\boldsymbol{X_1^T X_1}) \boldsymbol{b_1} + (\boldsymbol{X_1^T X_2}) \boldsymbol{b_2} = \boldsymbol{X_1^T y}$$

$$\boldsymbol{b_1}  = (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T y} - (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T X_2} \boldsymbol{b_2}$$

$$\boldsymbol{b_1}  = (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} (\boldsymbol{y}  - \boldsymbol{X_2} \boldsymbol{b_2})$$

The coefficients $\boldsymbol{b_1}$ are obtained from the regression of
$\boldsymbol{y}$ on $\boldsymbol{X_1}$ minus a correction vector

The regression of $\boldsymbol{y}$ on $\boldsymbol{X_1}$:

$$(\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{y}$$

The correction vector is:

$$(\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{X_2} \boldsymbol{b_2}$$

What is the **correction vector**?

The correction vector is the regression of the variables (done
separately and put together in a matrix) in $\boldsymbol{X_2}$ on all
the variables in $\boldsymbol{X_1}$.

The correction vector will be zero if and only if the variables in
$\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ are orthogonal.

The second term $\boldsymbol{b_2}$ will be zero if the variables in
$\boldsymbol{X_2}$ have no influence on $\boldsymbol{y}$, that is, the
variables in $\boldsymbol{X_2}$ are irrelevant to $\boldsymbol{y}$.

We can only omit $\boldsymbol{X_2}$ if the variables in
$\boldsymbol{X_2}$ are irrelevant.

If we omit relevant variables, then the parameter estimates in
$\boldsymbol{b_1}$ are affected.

If $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ are not orthogonal, then
the parameter estimates in $\boldsymbol{b_1}$ are affected.

## Deriving Estimates

What if $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ are not orthogonal?
Start with the partitioned normal equations:

$$\begin{matrix} (1) \\ (2) \end{matrix} \begin{bmatrix} \boldsymbol{X_1^T X_1} & \boldsymbol{X_1^T X_2} \\ \boldsymbol{X_2^T X_1} & \boldsymbol{X_2^T X_2} \end{bmatrix} \begin{bmatrix} \boldsymbol{b_1} \\ \boldsymbol{b_2} \end{bmatrix} = \begin{bmatrix} \boldsymbol{X_1^T y} \\ \boldsymbol{X_2^T y} \end{bmatrix}$$

From the partitioned normal equations, we can solve (1) for
$\boldsymbol{b_1}$:

$$\boldsymbol{b_1}  = (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} (\boldsymbol{y}  - \boldsymbol{X_2} \boldsymbol{b_2})$$

From the partitioned normal equations, we can use (2) to obtain:
$$(\boldsymbol{X_2^T X_1}) \boldsymbol{b_1} + (\boldsymbol{X_2^T X_2}) \boldsymbol{b_2} = \boldsymbol{X_2^T y}$$

Substituting $\boldsymbol{b_1}$ into the above yields:
$$(\boldsymbol{X_2^T X_1})  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} (\boldsymbol{y}  - \boldsymbol{X_2} \boldsymbol{b_2}) + (\boldsymbol{X_2^T X_2}) \boldsymbol{b_2} = \boldsymbol{X_2^T y}$$
$$(\boldsymbol{X_2^T X_1})  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{y}  - (\boldsymbol{X_2^T X_1})  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{X_2} \boldsymbol{b_2} + (\boldsymbol{X_2^T X_2}) \boldsymbol{b_2} = \boldsymbol{X_2^T y}$$

We can collect like terms to obtain:

$$\boldsymbol{X_2^T X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{y}  - \boldsymbol{X_2^T X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{X_2} \boldsymbol{b_2} + \boldsymbol{X_2^T X_2} \boldsymbol{b_2} = \boldsymbol{X_2^T y}$$

$$\boldsymbol{X_2^T X_2} \boldsymbol{b_2} - \boldsymbol{X_2^T X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{X_2} \boldsymbol{b_2} = \boldsymbol{X_2^T y} - \boldsymbol{X_2^T X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T} \boldsymbol{y}$$
$$[\boldsymbol{X_2^T}(\boldsymbol{I} - \boldsymbol{X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T}) \boldsymbol{X_2} ]\boldsymbol{b_2} = \boldsymbol{X_2^T}(\boldsymbol{I} - \boldsymbol{X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T}) \boldsymbol{y}$$

Now solve for $\boldsymbol{b_2}$
$$\boldsymbol{b_2} = [\boldsymbol{X_2^T}(\boldsymbol{I} - \boldsymbol{X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T}) \boldsymbol{X_2} ]^{-1}[\boldsymbol{X_2^T}(\boldsymbol{I} - \boldsymbol{X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T})\ \boldsymbol{y}]$$

## Residual Maker

Previously, we derived the **residual maker** matrix, **M**:

$$\boldsymbol{I - X(X^T X)^{-1}X} = \boldsymbol{M}$$

Remember that $\boldsymbol{M}$ is the residual maker, that is,
$\boldsymbol{My} = \boldsymbol{e}$.

In other words, $\boldsymbol{My}$ makes the residuals for the
regressions of $\boldsymbol{y}$ on $\boldsymbol{X}$.

The counterpart of **M** when for the partition $\boldsymbol{X_1}$ is:

$$\boldsymbol{I - X_1(X_1^T X_1)^{-1}X_1} = \boldsymbol{M_1}$$

This means that $\boldsymbol{M_1}$ is the residual maker for regressions
on the $\boldsymbol{X_1}$ variables.

$\boldsymbol{M_1 y}$ are the residuals from regressing $\boldsymbol{y}$
on the $\boldsymbol{X_1}$ variables.

$\boldsymbol{M_1 X_2}$ are the column by column residuals from
regressing each variable in $\boldsymbol{X_2}$ on all the variables in
$\boldsymbol{X_1}$.

We can use $\boldsymbol{M_1}$ to simplify:

$$\boldsymbol{b_2} = [\boldsymbol{X_2^T}(\boldsymbol{I} - \boldsymbol{X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T}) \boldsymbol{X_2} ]^{-1}[\boldsymbol{X_2^T}(\boldsymbol{I} - \boldsymbol{X_1}  (\boldsymbol{X_1^T X_1})^{-1} \boldsymbol{X_1^T})\ \boldsymbol{y}]$$

$$\boldsymbol{b_2} = [\boldsymbol{X_2^T}\ \boldsymbol{M_1}\ \boldsymbol{X_2} ]^{-1}[\boldsymbol{X_2^T}\ \boldsymbol{M_1}\ \boldsymbol{y}]$$

Since $\boldsymbol{M_1}$ is symmetric and idempotent,
$\boldsymbol{M_1} = \boldsymbol{M_1^T M_1}$.

$$\boldsymbol{b_2} = [\boldsymbol{X_2^T}\boldsymbol{M_1^T}\boldsymbol{M_1}  \boldsymbol{X_2} ]^{-1}[\boldsymbol{X_2^T}\boldsymbol{M_1^T}\boldsymbol{M_1} \boldsymbol{y}]$$

Let $\boldsymbol{M_1 X_2} = \boldsymbol{X_2^*}$ and
$\boldsymbol{M_1 y} = \boldsymbol{y^*}$

$$\boldsymbol{b_2} = [\boldsymbol{{X_2^*}^T}  \boldsymbol{X_2^*} ]^{-1}[\boldsymbol{{X_2^*}^T} \boldsymbol{y^*}]$$

## FWL Theorem

The Frisch-Waugh-Lovell Theorem discusses a fundamental result in
regression analysis.

In the linear least squares regression model of vector $\boldsymbol{y}$
on two sets of variables, $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$, the
subvector $\boldsymbol{b_2}$ is the set of coefficients obtained when
the residuals from a regression of $\boldsymbol{y}$ on
$\boldsymbol{X_1}$ alone are regressed on the set of residuals obtained
when each column of $\boldsymbol{X_2}$ is regressed on
$\boldsymbol{X_1}$.

The least squares coefficients $\boldsymbol{b_2}$ are obtained from
regressing $\boldsymbol{y^*}$ on $\boldsymbol{X_2^*}$.

$$\boldsymbol{b_2} = [\boldsymbol{{X_2^*}^T}  \boldsymbol{X_2^*} ]^{-1}[\boldsymbol{{X_2^*}^T} \boldsymbol{y^*}]$$

But what are $\boldsymbol{y^*}$ and $\boldsymbol{X_2^*}$?

$\boldsymbol{y^*}$ on $\boldsymbol{X_2^*}$ are just the residuals
obtained from regressing on the $\boldsymbol{X_1}$ variables.

The difference between regressing $\boldsymbol{y}$ only on
$\boldsymbol{X_2}$ and $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ is that
the latter first regresses the dependent variable and all the
$\boldsymbol{X_2}$ variables on the $\boldsymbol{X_1}$ variables and
then regresses the resulting residuals on each other.

The smaller regression just regresses $\boldsymbol{y}$ on the
$\boldsymbol{X_2}$ variables.

## What is the FWL Theorem?

The FWL theorem means that the multiple regression coefficient of any
single variable can be obtained by first netting out the effect of the
other variables in the regression model from the dependent variable and
the independent variables.

In practice this means taking the residuals from regressing
$\boldsymbol{y}$ on $\boldsymbol{X_1}$ and regressing these residuals on
the residuals from regressing $\boldsymbol{X_2}$ on $\boldsymbol{X_1}$.

This process is known as **partialing out** or **netting out** the
effect of $\boldsymbol{X_1}$.

In other words, a multiple regression model partials out the influence
of the the control variables.

Assume that our regression model was:

$$earnings = \beta_1 + \beta_2\ education + \beta_3\ age + \epsilon $$

We could first regress earnings on age and obtain the residuals. We
could then regress education on age and obtain the residuals.

Regressing the residuals for earnings on the residuals for education
would obtain the same coefficient as the full multiple regression model.

## FWL Example

In the example below, we first regress reading scores on the percentage
of students receiving free and reduced price lunch and average income in
the school district.

$$
reading_i = \beta_0 + \beta_1 lunch_i + \beta_2income_i + \epsilon_i
$$

We want to demonstrate the FWL Theorem. We do this by

-   Regressing reading scores on income

-   Obtaining the residuals from the regression of reading on income

-   Regressing the lunch variable on income

-   Obtaining the residuals from the regression of lunch on income

-   Regressing the residuals from the regression of **y** on **income**
    on the residuals from **lunch** on **income**

When we regress **y** on income, the resulting residuals are the
variation in **y** that is not explained by the income variable.

When we regress **lunch** on income, the resulting residuals are the
variation in **lunch** that are not explained by the income variable.

We then regress the unexplained variation in **y** on the unexplained
variation in **lunch**.

We have **netted out** the influence of **income**. The result should be
the partial coefficient that represents the marginal influence of lunch
on reading scores.

You can see in the table of regression estimates that the coefficient in
the full model for lunch is also the coefficient the the lunch residuals
in the regression of the **y residuals** on the **lunch residuals**.

As stated by the FWL Theorem, the coefficient obtained from the
regression of $\boldsymbol{M_1 y}$ on $\boldsymbol{M_1 X_2}$ is the same
as the coefficient from the multiple linear regression model on
$\boldsymbol{y}$ on $\boldsymbol{X}$.

This is why coefficients in multiple linear regression models are often
referred to as **partial regression coefficients.**

```{r, message = FALSE, warning = FALSE}

rm(list = ls())

library(AER)

data(CASchools)

# Multiple Regression

mod_1 <- lm(read ~ lunch + income, data = CASchools)

# Regress Y on X1
# Save residuals as m1_y

mod_2 <- lm(read ~ income, data = CASchools)
m1_y <- as.data.frame(mod_2$residuals)

# Regression X2 on X1
# Save residuals as m1_x1

mod_3 <- lm(lunch ~ income, data = CASchools)
m1_x1 <- as.data.frame(mod_3$residuals)

# Bind columns of residuals together in a dataframe
resids <- cbind(m1_y, m1_x1)

# Regress m1_y on m1_x1

fwl <- lm(mod_2$residuals ~ mod_3$residuals, data = resids)

modelsummary::modelsummary(list(mod_1, fwl))


```

## Summary

The Orthogonal Partitioned Regression theorem states that if you regress
$\boldsymbol{y}$ on two sets of variables, $\boldsymbol{X_1}$ and
$\boldsymbol{X_2}$, then if (and only if) $\boldsymbol{X_1}$ and
$\boldsymbol{X_2}$ are orthogonal, you can obtain the coefficient
vectors for $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$ by regressing
$\boldsymbol{y}$ on $\boldsymbol{X_1}$ and $\boldsymbol{X_2}$
separately.

What does this mean?

You can exclude the variables in $\boldsymbol{X_2}$ if these variables
are **irrelevant** to the dependent variable or **orthogonal** to the
other independent variables contained in $\boldsymbol{X_1}$.

If the variables in $\boldsymbol{X_2}$ are not irrelevant to the
dependent variable, then the exclusion of these variables will change
your estimates of $\boldsymbol{b_1}$.

If the variables in $\boldsymbol{X_2}$ are not orthogonal to the
variables in $\boldsymbol{X_1}$, then the exclusion of these variables
will change your estimates of $\boldsymbol{b_1}$.

## Orthogonal Example

In the R code, we create a large $\boldsymbol{X}$ matrix and estimate
$\boldsymbol{b}$.

Note that the elements of $\boldsymbol{X}$ are created randomly.

There is no underlying data generating process to suggest that the
elements of $\boldsymbol{X}$ violate the orthogonality assumption.

We partition $\boldsymbol{X}$ into $\boldsymbol{X_1}$ and
$\boldsymbol{X_2}$ and regress $\boldsymbol{y}$ on $\boldsymbol{X_1}$
and $\boldsymbol{X_2}$ separately.

Comparing the full and partitioned regression results for the
coefficients of the independent variables illustrates the partitioned
orthogonal regression theorem.

```{r}
rm(list = ls())
nrows <- 10000
ncols <- 7
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- rnorm(nrows)
        gc()
      }
      
beta <- c(1,2,3,4,5,6,7)

y <- x %*% beta + rnorm(nrows)

bhat <- solve(t(x) %*% x) %*% t(x) %*% y

x_1 <- x[,1:4]
x_2 <- cbind(x[,1],x[,5:7])

bhat_1 <- solve(t(x_1) %*% x_1) %*% t(x_1) %*% y
bhat_2 <- solve(t(x_2) %*% x_2) %*% t(x_2) %*% y

cbind(bhat[2:4],bhat_1[2:4])
cbind(bhat[5:7], bhat_2[2:4])

df <- data.frame(y = y, x = x)

```

## Non-Orthogonal Example

```{r, message = FALSE}

rm(list = ls())
library(MASS,quietly = TRUE)

nrows <- 10000
ncols <- 5

sigma_mat <- matrix(0, ncols-1, ncols-1) 
 sigma_mat[lower.tri(sigma_mat)] <- 0.95
 sigma_mat[upper.tri(sigma_mat)] <- 0.95
 diag(sigma_mat) <- 1

x <- matrix(0, nrows, ncols)
  x[,1] <- 1
  x[,2: ncols] <- mvrnorm(n = nrows, mu = c(1,2,3,4), 
                          Sigma = sigma_mat)

beta <- c(1,4,5,6,10)
y <- x %*% beta + rnorm(nrows)
bhat <- solve(t(x) %*% x) %*% t(x) %*% y

bhat
```

```{r}
x_1 <- x[,1:3]
x_2 <- cbind(x[,1],x[,4:5])

bhat_1 <- solve(t(x_1) %*% x_1) %*% t(x_1) %*% y
bhat_2 <- solve(t(x_2) %*% x_2) %*% t(x_2) %*% y

bhat_1
bhat_2

```

## Wrapping Up

If we omit independent variables that are not orthogonal to the other
independent variables, the least squares coefficients are adversely
impacted.

If we omit independent variables that are relevant to the dependent
variable, the least squares coefficients are adversely impacted.

The least squares coefficients from the multiple regression model are
partial regression coefficients.

A partial regression coefficient estimates the influence of the
independent variable after the influence of other independent variables
has been controlled for or netted out.

Hence, we preface our analysis with, "All else being equal" because we
have (hopefully) controlled for the influence of other relevant
variables of interest.
