---
title: "Heteroscedasticity"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

Module 6 focuses on the transition from parameter estimation to formal
statistical inference through the framework of hypothesis testing.
Students will examine the three fundamental approaches to
inference—Wald, fit-based, and Lagrange multiplier tests—to determine
whether empirical evidence supports specific theoretical restrictions on
population parameters. By mastering the application of $t$-tests for
individual coefficients, students will gain the analytical tools
necessary to assess the statistical significance of economic
relationships.

## Null Hypothesis

The **null hypothesis**, $H_0$, is the 'status quo' or typically the
assumption of 'no effect.'

The null hypothesis is the claim that we will 'operate under' unless
'strong evidence' otherwise.

In regression analysis, the null hypothesis is typically that the
regressor of interest has no impact on the dependent variable. In other
words,

$$
H_0: \beta_k = 0
$$

**We can only reject or fail to reject the null hypothesis given the
evidence at hand.**

In econometrics, we never "accept" the null hypothesis. We simply state
that the evidence is insufficient to disprove it.

## Alternative Hypothesis

The **alternative hypothesis** is the claim we are testing. Typically,
the alternative represents the presence of an effect, a difference, or a
relationship.

We do not test the alternative hypothesis. We test whether there is
sufficient evidence to reject the null hypothesis. In econometrics,
where the null hypothesis is often the true value of the population
parameter is equal to zero, we investigate whether, for a given sample,
there is sufficient evidence to reject the null of zero ('no effect').

In most cases, we have a two-sided null and alternative hypothesis, that
is, we are not testing a specific direction, merely whether we can
reject the null hypothesis of zero.

$$
H_A: \beta \ne 0
$$

## General Linear Hypotheses

We are imposing one or more linear restrictions on the linear
restriction model, that is, for the general linear model
$\boldsymbol{y = X\beta + \epsilon}$, we are imposing $\boldsymbol{J}$
linear restrictions.

Let $\boldsymbol{R}$ be a matrix where each row of $\boldsymbol{R}$
contains the coefficients of one of the restrictions.

$\boldsymbol{R}$ may have one or many rows.

We can thus write the $\boldsymbol{J}$ restrictions on
$\boldsymbol{\beta}$: $$\boldsymbol{R\ \beta\ =\ q}$$

The hypothesis implied by imposing the restrictions can be written as:
$$H_0:\ \boldsymbol{R\ \beta\ - \ q = 0}$$

$$H_A:\ \boldsymbol{R\ \beta\ - \ q \ne 0}$$

## One Linear Hypothesis

Assume that the regression model is:

$$
\boldsymbol{y = \beta_1 + \beta_2 x_2 + \beta_3 x_3 + \dots + \beta_k x_k + \epsilon}
$$

and we hypothesize that $\beta_3 = 0$ so that:

$$
H_0: \beta_3 = 0
$$

$$H_A: \beta_3 \ne 0$$

This means that the **R** matrix has 0's for every element EXCEPT for
$\beta_3$
$$\boldsymbol{R} = [0\quad 0\quad 1\quad 0\quad \dots\quad 0];\quad \boldsymbol{q} = 0$$

So, we can restate the null:

$$
H_0: \boldsymbol{R \beta - q = 0}
$$

$$H_0: \boldsymbol{R\beta} = [0\quad 0\quad 1\quad 0\quad \dots\quad 0]\begin{bmatrix}{\beta_1 \\ \beta_2 \\ \beta_3 \\ \vdots \\ \beta_k}\end{bmatrix}= 0 = \boldsymbol{q}$$

If we multiply this out, we find:

$$
H_0: (0 \times\beta_1)+(0 \times \beta_2)+(1 \times \beta_3) + \dots + (0 \times \beta_k) = 0
$$

$$
H_0: \beta_3 = 0
$$

## The T-Test

Previously, we derived the standard error for each estimated
coefficient.

We can test a single linear hypothesis using the **t-test**.

The t-test can be thought of as a 'ratio of distributions.'

For a single estimated coefficient, $b$, under the null hypothesis,
$H_0: \beta = 0$, the difference

$$
(b - \beta)
$$

measures the 'distance' from the estimate to the value of the population
coefficient under the null.

We normalize this 'distance' relative to the standard error of the
estimated coefficient.

The 'farther' the estimated coefficient is from the hypothesized value
under the null, expressed in standard error units, the more likely the
estimated coefficient is not equal to the hypothesized value under the
null.

The t-test is equal to:

$$
t_{n-K} \sim \frac{b - \beta}{se(b)}
$$

## Decision Rule: T-test

To decide whether to reject $H_0$, we compare the calculated t-statistic
to a **critical value**.

We use different levels of statistical significance, ranging from
$\alpha = 0.10$ (10% statistical significance) to $\alpha = 0.05$ (5%
statistical significance) to $\alpha = 0.01$ (1% level of statistical
significance).

If the absolute value of the calculated t-statistic is greater than the
critical value, then we reject the null hypothesis. On the other hand,
if the absolute value of the t-statistic is less than or equal to the
critical value of the t-statistic, we fail to reject the null hypothesis
or:

$$
|t_{calc}| > t_{crit} \, \rightarrow \text{reject} \, H_0
$$

$$
|t_{calc}| \le t_{crit} \, \rightarrow \text{fail to reject} \, H_0
$$

## Example in R

In the code chunk below, we demonstrate how to estimate the t-statistic
for simulated data and test the null hypothesis that each coefficient is
equal to zero.

```{r}

rm(list = ls())
set.seed(1234)

nrows <- 50

x <- cbind(1, 
           runif(nrows, 0, 20), 
           runif(nrows, 0, 20),
           rnorm(nrows, 5, 3)) 

beta <- c(10, 2, -1, 3)
epsilon <- rnorm(nrows, mean = 0, sd = 3.75)

y <- x %*% beta + epsilon

bhat <- solve(t(x) %*% x) %*% t(x) %*% y

e <- y - x %*% bhat
n <- nrow(x)
k <- ncol(x)
s2 <- as.numeric((t(e) %*% e) / (n - k))

var_cov_b <- s2 * solve(t(x) %*% x)
se_b <- sqrt(diag(var_cov_b))

# 5. T-test (Two-Tailed)
beta <- 0

t_stat <- (bhat - beta) / se_b

p_values <- 2 * pt(abs(t_stat), df = n - k, lower.tail = FALSE)

df <- tibble(round(bhat,3),
             round(se_b,3),
             round(t_stat,3),
             round(p_values,5))

kable(df,
      align     = 'c',
      digits    = 3,
      col.names = c('Beta Hat', 'Standard Error',
                    'T Statistic', 'P-Value')) %>%
kable_classic()


```
