---
title: "Deriving Least Squares"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Least Squares

In this section, we discuss the motivation behind least squares regression before deriving the least squares estimator. We then examine the small sample properties of the least squares estimator. We explore how the dependent variable can be decomposed into **explained** and **unexplained** variation through the use of the **residual maker** and **projection** matrices.

## Notation

Define a column vector $\boldsymbol{x_k}$ such that $\boldsymbol{x_k}$ consists of $n$ observations.

Let $k \in \{1, 2, \dots, K\}$ such that $\boldsymbol{x_1}$ is the first column vector.

Let $\boldsymbol{x_2}$ be the second column vector, and so on.

Combining the column vectors for $\boldsymbol{x_k}$ yields a $(n \times K)$ matrix.

The first column of the resulting $\boldsymbol{X}$ matrix consists of a column of 1's.

The subscript $k$ refers to the k-th column of the $\boldsymbol{X}$ matrix.

The subscript $i$ refers to the i-th row of the $\boldsymbol{X}$ matrix.

The linear regression model is equal to:

$$\boldsymbol{y = X \, \beta + \epsilon}$$

$$\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_n
\end{bmatrix} =
\begin{bmatrix}
x_{11} & x_{12} & \dots & x_{1K} \\
x_{21} & x_{22} & \dots & x_{2K} \\
\vdots & \vdots & \vdots & \vdots \\
x_{n1} & x_{n2} & x_{n3} & x_{nK} 
\end{bmatrix}
\begin{bmatrix}
\beta_1 \\
\beta_2 \\
\beta_3 \\
\beta_4
\end{bmatrix} +
\begin{bmatrix}
\epsilon_1 \\
\epsilon_2 \\
\vdots \\
\epsilon_n
\end{bmatrix}$$

## Why Least Squares?

Assume that we wish to estimate the impact of education on workers' salaries. We specify a linear regression model where salary is a function of education and other explanatory variables or:

$$
salary = \beta_1 + \beta_2 \, education + \beta_3 \, age + \dots + \beta_k x_K + \epsilon
$$

We do not observe the **population parameters**.

We need an **algorithm** or **rule** by which we can develop **estimates** of the population parameters.

The algorithm or rule is called an **estimator**.

When we use the estimator with data, we produce estimates of the population parameters.

***What is our objective?***

We want an algorithm or rule that produces estimates that are "close" to the actual values.

In other words, we want an estimator to produce estimates that are "close" to the actual values of the dependent variable.

We can define our estimates of the population parameters as **b**.

We can define our predicted values of the dependent variable as $\hat{y}$.

Our predicted values are equal to:

$$
\hat{y} = Xb = b_1x_1 + b_2x_2 + \dots + b_k x_k
$$

And our "mistakes" or "residuals" are defined as:

$$
e = y - \hat{y} = y - Xb
$$

So, what is our **objective function**?

We could try and minimize the sum of our mistakes or:

$$
\sum_{i=1}^n = y_i - \hat{y_i} = \sum_{i = 1}^n = e_i
$$

***However, we assume that our errors are normally distributed with a mean of zero, so if we sum the residuals, we would add the positive and negative values together and they would be equal to zero.***

$$
\sum_{i=1}^n  y_i - \hat{y_i} = \sum_{i = 1}^n  e_i = 0
$$

Instead, we can square our residuals and sum the squared residuals.

This is similar to estimating the variance of a random variable where we sum the squared deviations of the variable from its mean and then divide by $N$.

$$
var(X) = \frac{1}{N}\sum_{i=1}^N (x_i - \mu_x)^2
$$

Our objective function is **the minimization of the sum of the squared residuals:**

$$
\sum_{i = 1}^n (y_i - \hat{y})^2  = \sum_{i=1}^n e_i^2
$$

The question is: how do we arrive that estimates of the population parameters that minimize the sum of the squared residuals?

## Objective Function

Let $\boldsymbol{b_0}$ be a coefficient vector consisting of sample estimates of $\boldsymbol{\beta}$ from a least squares regression.

The population regression function is:

$$\boldsymbol{y} = \boldsymbol{X\beta} + \boldsymbol{\epsilon}$$

The sample regression function is:

$$\boldsymbol{y} = \boldsymbol{X\ b_0} + \boldsymbol{e}$$

where the predicted values of the dependent variable are equal to:

$$
\boldsymbol{\hat{y}} = \boldsymbol{X \, b_o}
$$

and the residuals are equal to:

$$
\boldsymbol{e} = \boldsymbol{y - \hat{y}} = \boldsymbol{y - X \, b_0}
$$

Our objective function is the **minimization of the sum of the squared residuals**, that is, we choose $\boldsymbol{b_0}$ such that the predicted values of the dependent variable best "fit" the data.

The sum of the squared residuals when we select $\boldsymbol{b_0}$ is equal to:

$$
\boldsymbol{e_0^T \, e_0}
$$

We can now specific the objective function as:

$$min_{b0}\ S(b_0) = \boldsymbol{e_0^T e_0} = (\boldsymbol{y - \hat{y}})^T\  (\boldsymbol{y - \hat{y}})$$

$$e_0^T e_0 = (\boldsymbol{y - X\ b_0})^T\  (\boldsymbol{y - X\ b_0})$$

$$e_0^T e_0 = \boldsymbol{y^Ty - b_0^T\ X^T y - y^T\ Xb_0 + b_0^T X^T X b_0}$$ It is important to note that:

$$\boldsymbol{b_0^T X^T y} \sim (1 \times K) \times (K \times n) \times (n \times 1) \sim (1 \times 1)$$ $$\boldsymbol{y^T X b_0} \sim (1 \times n) \times (n \times K) \times (K \times 1) \sim (1 \times 1)$$

$$(\boldsymbol{b_0^T\ X^T y})^T = \boldsymbol{y^T\ X\ b_0}$$

We can substitute for $\boldsymbol{y^T\ X\ b_0}$ to obtain:

$$min_{b0}\ S(b_0) =e_0^T e_0 = \boldsymbol{y^Ty - 2b_0^T\ X^T\ y + b_0^T X^T X b_0}$$

## First Derivative

There are two useful identities that will help us in the derivation of the least squares estimator.

First, for the vectors $\boldsymbol{a}$ and $\boldsymbol{b}$, the partial derivative of $\boldsymbol{a^T}$ times $\boldsymbol{b}$ with respect to $\boldsymbol{a}$ is equal to $\boldsymbol{b}$.

$$\frac{\delta \boldsymbol{a^Tb}}{\delta \boldsymbol{a}} = \boldsymbol{b}$$

Second, if $\boldsymbol{A}$ is symmetric, then the first derivative of $\boldsymbol{b^T A b}$ with respect to $\boldsymbol{b}$ is:

$$\frac{\delta \boldsymbol{b^T A b}}{\delta \boldsymbol{b}} = 2 \boldsymbol{Ab} = 2 \boldsymbol{b^TA}$$

We can now return to the task at hand. We need to take the first derivative of the objective function with respect to $\boldsymbol{b_0}$ and set the first derivative equal to zero or:

$$\frac{\partial(\boldsymbol{e_0^T e_0 )}}{\partial\ \boldsymbol{b_0}} = 0$$

$$\frac{\partial(\boldsymbol{e_0^T e_0 )}}{\partial\ \boldsymbol{b_0}} =\frac{\partial(\boldsymbol{y^Ty - 2b_0^T\ X^T\ y + b_0^T X^T X b_0})}{\boldsymbol{b_0}} = 0$$

$$\frac{\partial(e_0^T e_0 )}{\partial\ b_0} =\frac{\partial \,y^Ty}{\partial{b_0}} +  \frac{\partial \, 2b_0^T\ X^T\ y}{\partial b_0} + \frac{\partial \, b_0^T X^T Xb_0}{\partial \, b_0} = 0$$

Breaking the first derivative into its component parts yields:

$$\frac{\partial \,y^Ty}{\partial{b_0}} = 0 $$

$$\frac{\partial\ (-2\ \boldsymbol{b_0^T X^T y)}}{\partial\ \boldsymbol{b_0}} = -2 \boldsymbol{X^T y}$$ $$\frac{\partial\ (\boldsymbol{b_0^T X^T X b_0})}{\partial\ \boldsymbol{b_0}} = 2\ \boldsymbol{X^T X\ b_0}$$

The first derivative of the objective function is equal to: $$\frac{\partial S(\boldsymbol{b_0})}{\partial \boldsymbol{b_0}} = -2\ \boldsymbol{X^T\ y} + 2\ \boldsymbol{X^T X\ b_0} = 0$$

If a solution $\boldsymbol{b}$ exists, then the **normal equations** are: $$\boldsymbol{X^T\ X\ b} = \boldsymbol{X^T\ y}$$

## Least Squares Estimator

If the full column rank assumption holds, then $(\boldsymbol{X^T X})$ is square as it is $(K \times K)$ and $(\boldsymbol{X^T X})$ is also symmetric.

If the full column rank assumption holds, then $(\boldsymbol{X^T X})$ has an inverse and we can use the normal equations to obtain the least square estimates of the population parameters.

$$\boldsymbol{X^T\ X\ b} = \boldsymbol{X^T\ y}$$

Given that the inverse of $\boldsymbol{(X^T X)}$ exists, we can now solve for the **least squares estimator**.

We define the least squares estimator as "**beta hat**" and use **b** to denote the least squares estimator.

$$\boldsymbol{b} = (\boldsymbol{X^T\ X})^{-1}\ \boldsymbol{X^T\ y}$$

## Second Derivative

If the first derivative is equal to zero, it informs us that the objective function has arrived at a minimum (or maximum). The first derivative being zero only means we have arrived at a stationary point.

**We need the second derivative to provide that we have found a minimum instead of a maximum or a saddle point**.

Starting with the first derivative:

$$\frac{\partial S(\boldsymbol{b_0})}{\partial \boldsymbol{b_0}} = -2\ \boldsymbol{X^T\ y} + 2\ \boldsymbol{X^T X\ b_0} = 0$$

We take the derivative of the first derivative to obtain the second derivative : $$\frac{\partial^2 S(\boldsymbol{b_0})}{\partial \boldsymbol{b_0 b_0^T}} = \frac{\partial}{\partial \boldsymbol{b_0 }} (-2\ \boldsymbol{X^T\ y} + 2\ \boldsymbol{X^T X\ b_0}) $$

We know that:

$$\frac{\partial}{\partial \boldsymbol{b_0}} (-2 \boldsymbol{X^T\ y}) = 0$$ $$\frac{\partial}{\partial \boldsymbol{b_0}} (-2 \boldsymbol{X^T X b_0}) = \boldsymbol{2 X^T X}$$

In other words, the second derivative is equal to:

$$\frac{\partial^2 S(\boldsymbol{b_0})}{\partial \boldsymbol{b_0 b_0^T}} = 2\ \boldsymbol{X^T X}$$

For a point to be a **local minimum**, the Hessian matrix (second derivative) must be **Positive Definite.**

We know that $\boldsymbol{X^TX}$ is always symmetric

We know that, for any non-zero vector, the quadratic form is at least positive semi-definite.

$$
\boldsymbol{v^T (X^TX)} = \boldsymbol{(Xv)^T(Xv)} = \boldsymbol{||Xv||^2} 
$$

We know that, if $\boldsymbol{X}$ has full column rank, then $\boldsymbol{||Xv||^2} > 0$, which means that the second derivative is positive definite.

**Because the second derivative is positive definite, the objective function is strictly convex and the stationary point (b) is a unique global minimum.\
\
**Simply put, because of the assumptions of linearity, full rank, and zero conditional mean, the least squares estimator minimizes (globally) the sum of the squared residuals.

## Orthogonality of X and e

We can use $\boldsymbol{y} = \boldsymbol{X\ b} + \boldsymbol{e}$ and the normal equations to obtain: $$\boldsymbol{X^T\ X\ b} = \boldsymbol{X^T\ y}$$ $$\boldsymbol{X^T X\ b} = \boldsymbol{X^T\ (\boldsymbol{X\ b} + \boldsymbol{e})}$$ $$\boldsymbol{X^T X\ b} = (\boldsymbol{X^T X}) \boldsymbol{b} + \boldsymbol{X^T e}$$ $$\boldsymbol{X^T e} = 0$$

Note that $\boldsymbol{X^T e} = 0$ means that $\boldsymbol{x_k^T e} = 0$.

The covariance between each independent variable and the disturbances is equal to zero.

This only occurs if the assumption of the exogeneity of the regressors (**zero conditional mean**) holds.

This illustrates why the assumptions of linearity, full rank, and strict exogeneity of the regressors are necessary to obtain the least squares estimates.

## Projection

Imagine you have a flashlight and you are standing in a room.

The **floor** represents all the possible values that can be created by your **independent** variables.

There is a **balloon** floating above the floor. Let the balloon represent the dependent variable, **y**. Note that unless your model is perfect, the balloon will never touch the floor.\
\
**In other words, even if we have all the independent variables, we cannot perfectly explain y**.

If you shine a flashlight directly above the balloon, you get a **shadow** on the floor. The shadow presents the predicted value of **y** created by the independent variables or $\boldsymbol{\hat{y}}$.

In other words, the predicted value of **y** on the floor is the projection of **y** on **X**.

***The projection of y on X is the best linear approximation of y using the information in X.***

What is the best projection of **y** on **X**? Think of the lines from the floor to the balloon. We have a line at a 10 degree angle, 25 degree angle, and so on until we have a line at a 90 degree angle from the floor to the balloon.\
\
We want the linear that minimizes the distance from our predicted value of y to the actual value of y.

A straight line from the floor to the balloon minimizes the distance, in other words, a line at a 90 degree angle from the floor, that is, a **perpendicular line from y onto the floor**.

Why is the "perpendicular line" the same as "minimizing the sum of squares"?

Think of the distance between the balloon, $\boldsymbol{y}$, and any point on the floor, $\boldsymbol{\hat{y}}$.

By the **Pythagorean Theorem**, the shortest distance between a point and a plane is always the **perpendicular** line.

The length of that perpendicular line is the magnitude of our residuals, $\boldsymbol{||e||}$.

Minimizing the squared length of the residuals is mathematically identical to finding the orthogonal projection.

We now need to define **how** we can make the projection happen and what is "left over."

## Projection Matrix

We have previously defined our linear regression model as:

$$
\boldsymbol{y = X\beta + \epsilon = Xb + e}
$$

We know that our predicted values are equal to:

$$
\boldsymbol{\hat{y} = Xb = y - e}
$$

We can use this to object the **projection matrix** $\boldsymbol{P}$ $$\boldsymbol{\hat{y}} = \boldsymbol{Xb}$$ $$\boldsymbol{\hat{y}} = \boldsymbol{X(X^TX)^{-1}X^T y}$$\
$$\boldsymbol{\hat{y}} = \boldsymbol{X (X^T X)^{-1} X^Ty} = \boldsymbol{P y}$$

$$\boldsymbol{P} = \boldsymbol{X (X^T X)^{-1} X^T}$$

The projection matrix $\boldsymbol{P}$ is symmetric and idempotent.

## Residual Maker Matrix

The vector of least square residuals is $\boldsymbol{e} = \boldsymbol{y} - \boldsymbol{X b}$.

Substituting for $\boldsymbol{b}$ yields:

$$\boldsymbol{e} = \boldsymbol{y} - \boldsymbol{X b}$$ $$\boldsymbol{e} = \boldsymbol{y} - \boldsymbol{X (X^T X)^{-1} X^T y}$$ $$\boldsymbol{e} = (\boldsymbol{I} - \boldsymbol{X (X^T X)^{-1} X^T})\boldsymbol{y}$$ $$\boldsymbol{e} = \boldsymbol{M y}$$

$$\boldsymbol{M} = (\boldsymbol{I} - \boldsymbol{X (X^T X)^{-1} X^T})$$

**M** is defined as the **residual maker** or the annihilator **matrix**.

## M and P

**M** is symmetric and idempotent.

**P** is symmetric and idempotent.

**M and P are orthogonal.**

The residual maker and the projection matrix are orthogonal. $$\boldsymbol{PM} = \boldsymbol{MP} = \boldsymbol{0}$$

$$\boldsymbol{PM} = (\boldsymbol{X (X^T X)^{-1} X^T})(\boldsymbol{I} - \boldsymbol{X (X^T X)^{-1} X^T})$$ $$\boldsymbol{PM} = \boldsymbol{X (X^T X)^{-1} X^T} - \boldsymbol{X (X^T X)^{-1} X^T}\boldsymbol{X (X^T X)^{-1} X^T}$$ $$\boldsymbol{PM} = \boldsymbol{X (X^T X)^{-1} X^T} - \boldsymbol{X (X^T X)^{-1} X^T} = \boldsymbol{0}$$

When $\boldsymbol{y}$ is pre-multiplied by $\boldsymbol{M}$, it produces a vector of the least squares residuals:

$$\boldsymbol{e} = \boldsymbol{M y}$$

When $\boldsymbol{X}$ is pre-multiplied by $\boldsymbol{M}$ the result is: $$\boldsymbol{MX} = (\boldsymbol{I} - \boldsymbol{X (X^T X)^{-1} X^T})X$$ $$\boldsymbol{MX} = \boldsymbol{X} - \boldsymbol{X (X^T X)^{-1} (X^T X)} = \boldsymbol{(X - X)} = \boldsymbol{0}$$

The regression of $\boldsymbol{X}$ on $\boldsymbol{X}$ is a perfect fit and the residuals from this regression are equal to 0.

## Decomposing Y

We can now partition $\boldsymbol{y}$ into two orthogonal parts, the **projection** and the **residual** $$\boldsymbol{y} = \boldsymbol{Py} + \boldsymbol{My}$$

Furthermore, the sum of squares $\boldsymbol{y^T y}$ exploits the orthogonality between $\boldsymbol{P}$ and $\boldsymbol{M}$: $$\boldsymbol{y^T y} = \boldsymbol{y^T P^T P y} + \boldsymbol{y^T M^T M y}$$ $$\boldsymbol{y^T y} = \boldsymbol{\hat{y}^T \hat{y}} + \boldsymbol{e^T e}$$ The projection of the vector $\boldsymbol{y}$ into the column space of $\boldsymbol{X}$ creates fitted values $\boldsymbol{\hat{y}}$ and residuals $\boldsymbol{e}$.

In other words, the linear least squares regression projects the dependent variable into the independent variable space, but does not do so "perfectly."

The residual vector $\boldsymbol{e}$ captures the differences between the actual and predicted values of $\boldsymbol{y}$.

If the linearity assumption holds, then the population conditional mean $E[y | \boldsymbol{x}]$ is linear in $\boldsymbol{x}$.

The population conditional mean is the population counterpart to the projection of $y$ on $\boldsymbol{x}$.

```{r}

rm(list = ls())

X <- matrix(c(1,2,3,4,
              4,16,27,38,
              19,10,31,2,
              43,54,65,96), 
              nrow = 4, ncol = 4)

y <- c(6, 7, 8, 9) 

I <- diag(4)       

# Define P and M

P <- X %*% solve(t(X) %*% X) %*% t(X)
M <- I - X %*% solve(t(X) %*% X) %*% t(X)

y_hat <- P*y
e <- M*y

# Show P and M orthogonal

pm <- P %*% M

# Show MX = 0

mx <- M %*% X


```
