---
title: "Orthogonality"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

This module extends our discussion of **orthogonality**.

Orthogonality serves as the geometric bedrock of linear regression, providing a rigorous mathematical framework for the concept of "unrelatedness" between variables. At its simplest level, orthogonality is defined by a **zero dot product** between two vectors, indicating that they are perpendicular and share no common components in a vector space.

In the context of least squares, this principle allows us to decompose the dependent variable $\boldsymbol{y}$ into two perfectly independent parts: the projection, $\boldsymbol{\hat{y}}$, which represents the information captured by the model, and the residual vector $\boldsymbol{e}$, which represents the remaining noise. By ensuring that the residuals are orthogonal to the column space of the regressors, the least squares estimator effectively "annihilates" all linear information contained in the independent variables from the error term, resulting in a unique, global minimum where the predicted signal and the unexplained noise are completely uncorrelated.

## Dot Product

The **dot product** or **inner product** or **scalar product** is a way to multiply two vectors to get a scalar. The dot product measures how much the two vectors are aligned.

Given two vectors, **a** and **b**, where $a = (a_1, a_2, \dots, a_n)$ and $b = (b_1, b_2, \dots, b_n)$, the dot product of **a** and **b** is equal to:

$$
a \cdot b = a_1 b_1 + a_2 b_2 + a_3 b_3 + \dots + a_n b_n
$$

## Orthogonality

If two vectors are orthogonal, then the two vectors are **perpendicular** to each other.

Let $\theta$ be the angle between vectors **a** and **b**. The dot product of the two vectors is

$$
a \cdot b = |a| |b| cos(\theta)
$$

If the two vectors are orthogonal, then the angle between the vectors is 90 degrees.

The cosine of 90 degrees is equal to 0, thus the dot product between the two vectors is equal to 0. This means that the two vectors have no component in the same direction.

## Orthogonal Example

Assume that we have two vectors,

$$
a = (3,4) \quad \quad b = (6,8)
$$

The dot product between **a** and **b** is equal to:

$$
a \cdot b = (3 \times 6) + (4 \times 8) = 18 + 32 = 50
$$

In this example, the two vectors are **not orthogonal** to each other.

Assume we have two vectors,

$$
a = (3,4) \quad \quad c = (-4,3)
$$

The dot product between **a** and **c** is equal to:

$$
a \cdot c = (3 \times -4) + (4 \times 3) = -12 + 12 = 0
$$

In this example, **a** and **c** are orthogonal as their dot product is equal to 0.

In the code example below, we show that

$$
a \cdot b = 50 \\
a \cdot c = 0 \\
b \cdot c = 0
$$

```{r example}

# Define vectors

vector1 <- c(3, 4)
vector2 <- c(6, 8)  
vector3 <- c(-4,3) 

# Check whether dot product is equal to 0

dot_prod_12 <- vector1 %*% vector2
dot_prod_13 <- vector1 %*% vector3
dot_prod_23 <- vector2 %*% vector3

# Create a data frame for plotting
df <- tibble(x = c(0, 0), 
             y = c(0, 0), 
             xend = c(vector1[1], vector3[1]), 
             yend = c(vector1[2], vector3[2]))

# Plot the vectors
ggplot() +
  geom_segment(data = df, 
               aes(x = x, 
                   y = y, 
                   xend = xend, 
                   yend = yend),
               arrow = arrow(length = unit(0.2, "inches")),
               color = c("blue", "red"), linewidth = 1) +
  xlim(-5, 5) + ylim(-5, 5) +
  coord_fixed() +  
  theme_minimal() +
  ggtitle("Orthogonal Vectors in R") +
  xlab("X-axis") + 
  ylab("Y-axis") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed")

```

## Least Squares

Given the assumptions of linearity, full rank, and zero conditional mean, the least squares estimator is equal to:

$$
b = (X^T X)^{-1} X^T y
$$

The predicted (fitted) values, $\boldsymbol{\hat{y}}$ are equal to

$$
\boldsymbol{\hat{y} = Xb}
$$

The residuals are thus equal to

$$
e = y - \hat{y} = y - Xb = y - X(X^TX)^{-1}X^Ty = My
$$

We know that **M** and **P** are orthogonal, so it stands to reason that **My** and **Py** are orthogonal, or

$$
(My)^T Py = 0
$$

$$
(My)^T Py = ((I - P)y)^T Py = y^T(I-P)Py
$$

Recalling that **I** and **P** are square, symmetric, and idempotent yields

$$
(My)^T Py = y^T(I - P)Py = y^T(P-P^2)y = y^T(P-P)y = 0
$$

Thus, we have demonstrated that **My** and **Py** are orthogonal.

We can demonstrate this below using R.

Here we use the **mtcars** example data to create a **y** vector containing *mpg* and a matrix of independent variables where the first column is a column of 1's.

We estimate **b** and **Xb** to obtain **e** and then illustrate the dot product of $e^T$ and $\hat{y}$ is approximately 0.

We then turn to the question of whether $\boldsymbol{e}$ is orthogonal to each individual regressor, that is, does zero conditional mean hold? The answer is yes, the residuals are orthogonal to the regressors.

```{r example2}

rm(list = ls())

data(mtcars)

y <- as.matrix(mtcars$mpg)

x <- as.matrix(tibble(x1 = rep(1,32),
            cyl = mtcars$cyl,
            hp  = mtcars$hp,
            wt  = mtcars$wt))

I <- diag(32)

P <- x %*% solve(t(x) %*% x) %*% t(x)

M <- (I - P)

bhat <- solve(t(x) %*% x) %*% t(x) %*% y

e <- M %*% y

yhat <- x %*% bhat

dot_e_yhat <- t(e) %*% yhat

orth_e_cyl <- t(e) %*% x[,"cyl"]
orth_e_hp  <- t(e) %*% x[,"hp"]
orth_e_wt  <- t(e) %*% x[,"wt"]

df <- tibble(round(dot_e_yhat, 3),
             round(orth_e_cyl,3),
             round(orth_e_hp,3),
             round(orth_e_wt,3))

kable(df,
      col.names = c('Dot Product of Yhat and e',
                    'Dot Product of e and Cylinder Variable',
                    'Dot Product of e and HP Variable',
                    'Dot Product of e and Weight Variable'),
      align     = 'c') %>%
kable_styling()

```
