---
title: "Matrix Operations"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Matrix Operations

This module covers the essential matrix operations required for econometrics. We begin with basic matrix creation and arithmetic, progressing to more complex operations like inversion and formal multiplication. Understanding these operations is critical as they form the backbone of the Ordinary Least Squares (OLS) estimator, which we will derive and implement both manually and using built-in R functions.

## Matrix Creation

We can create two vectors that contain 4 numeric elements and form two matrices that are $(2 \times 2)$ Assume that we have call the two matrices, $m_1$ and $m_2$ and the matrices can be expressed as:

$$ m_1 = \begin{bmatrix} 2 & 6 \\ 4 & 8 \end{bmatrix} $$

$$ m_2 = \begin{bmatrix} 5 & 2 \\ 15 & 12 \end{bmatrix} $$

We can also create a $(4 \times 1)$ matrix from the same elements that form $m_1$ or:

$$
m_{1a} = \begin{bmatrix}
2 \\
4 \\ 
6 \\ 
8
\end{bmatrix}
$$

We can take a transpose of a matrix by using **t()**, so **t(m_1)** would be the transpose of matrix *m_1* or

$$
m_1^T = \begin{bmatrix}
2 & 4 \\
6 & 8
\end{bmatrix}
$$

We can use the **dim** function to obtain the dimensions of a matrix (and other elements like vectors).

The code chunk below does the following:

1.  Create numeric vectors using the **combine** function and assign to **d_1** and **d_2**

2.  Using **d_1**, create a $(2 \times 2)$ matrix

3.  Using **d_2**, create a $(2 \times 2)$ matrix

4.  Using **d_1**, create a \$(4 \times 1) matrix

5.  Use **kable** to output **m_1**

6.  Use the **transpose** function and **kable** to output **t(m_1)**

```{webr-r}

rm(list = ls())

d_1 <- c(2, 4, 6, 8)
d_2 <- c(5, 15, 2, 12)

m_1 <- matrix(d_1, nrow = 2, ncol = 2)
m_2 <- matrix(d_2, nrow = 2, ncol = 2)

m_1_a <- matrix(d_1, ncol = 4, nrow = 1)

dims <- tibble(d_m1     = dim(m_1),
               d_m2     = dim(m_2),
               d_m1_a   = dim(m_1_a),
               d_t_m1_a = dim(t(m_1_a)))

kable(m_1,
      caption = "Matrix M1") 

kable(t(m_1),
      caption = "Transpose of Matrix M1") 

kable(dims,
      col.names = c('M1', 'M2', 'M_1_A', 'T(M_1_A)'),
      full_width = TRUE)
```

## Matrix Addition

In this example, we demonstrate how to add matrices.

We return to the two $(2 \times 2)$ matrices in the previous example and add the matrices together.

$$ \begin{bmatrix} 
2 & 6 \\ 4 & 8 
\end{bmatrix} + 
\begin{bmatrix}
5 & 2 \\ 15 & 12
\end{bmatrix} =
\begin{bmatrix}
(2 + 5) & (6 + 2) \\
(4 + 15) & (8 + 12)
\end{bmatrix} = 
\begin{bmatrix}
7 & 8 \\
19 & 20 
\end{bmatrix}$$

When we add $m_1$ and $m_2$, we are adding the values in the same position in each matrix.

For example, for $m_1$, the value in the first row, first column is equal to 2.

For $m_2$, the value in the first row, first column is equal to 5.

Adding $m_1$ and $m_2$ yields a value of 7 in the first row, first column of the resulting matrix.

*The order for matrix addition does not matter, that is,* $m_1 + m_2 = m_2 + m_1$.

```{webr-r}

rm(list = ls())

d_1 <- c(2, 4, 6, 8)
d_2 <- c(5, 15, 2, 12)

m_1 <- matrix(d_1, nrow = 2, ncol = 2)
m_2 <- matrix(d_2, nrow = 2, ncol = 2)

# Matrix Addition

m_plus_1_2 <- m_1 + m_2

kable(m_plus_1_2,
      caption = "Element by Element Addition (M1+M2)")

```

## Matrix Subtraction

Using a similar process, we can subtract one matrix from another. While order does not matter for addition, order does matter for subtraction.

If we subtract $m_2$ from $m_1$, that is $m_1 - m_2$, then, we are subtracting the value of each element in $m_2$ from the corresponding element in $m_1$.

$$ \begin{bmatrix} 
2 & 6 \\ 4 & 8 
\end{bmatrix} - 
\begin{bmatrix}
5 & 2 \\ 15 & 12
\end{bmatrix} =
\begin{bmatrix}
(2 - 5) & (6 - 2) \\
(4 - 15) & (8 - 12)
\end{bmatrix} = 
\begin{bmatrix}
-3 & 4 \\
-11 & -4 
\end{bmatrix}$$

The value in the first row, first column of $m_1$ is equal to 5. When we subtract $m_2$ from $m_1$, we would subtract 2 (first row, first column of $m_2$) from 5 (first row, first column of $m_1$), that is, the resulting value is equal to $2-5 = -3$.

On the other hand, if we subtract $m_1$ from $m_2$, then the same position yields $5-2 = 3$.

$$ \begin{bmatrix}
5 & 2 \\ 15 & 12
\end{bmatrix} -
\begin{bmatrix} 
2 & 6 \\ 4 & 8 
\end{bmatrix} - 
 =
\begin{bmatrix}
(5 - 2) & (2 - 6) \\
(15 - 4) & (12 - 8)
\end{bmatrix} = 
\begin{bmatrix}
3 & -4 \\
11 & 4 
\end{bmatrix}$$

```{webr-r}

rm(list = ls())

d_1 <- c(2, 4, 6, 8)
d_2 <- c(5, 15, 2, 12)

m_1 <- matrix(d_1, nrow = 2, ncol = 2)
m_2 <- matrix(d_2, nrow = 2, ncol = 2)

# Matrix Subtraction

m_minus_1_2 <- m_1 - m_2

kable(m_minus_1_2,
      caption = "Element by Element Subtraction (M1-M2)") 

# Matrix Subtraction 

m_minus_2_1 <- m_2 - m_1

kable(m_minus_2_1,
      caption = "Element by Element Subtraction (M2-M1)") 
```

## Hadamard Product

Element-by-element matrix multiplication is straightforward but is not typically used in this class. Element-by-element matrix multiplication is known as the *Hadamard Product*.

The order of multiplication does not matter, however, the two matrices must have **identical dimensions**.

Returning to the two matrices, $m_1$ and $m_2$, we can conduct element-by-element multiplication by multiplying each element by its corresponding element in the other matrix.

$$
m1 \times m2 = 
\begin{bmatrix}
2 & 6 \\
4 & 8
\end{bmatrix} \times
\begin{bmatrix}
5 & 2 \\
15 & 12 
\end{bmatrix} =
\begin{bmatrix}
(2 \times 5) & (6 \times 2) \\
(4 \times 15) & (8 \times 12)
\end{bmatrix} =
\begin{bmatrix}
10 & 12 \\
60 & 96
\end{bmatrix}
$$

As an exercise, you can use the **webr** code chunk below to calculate the product of the two matrices and observe whether the Hadamard Product depends on the order of multiplication.

```{webr-r}

rm(list = ls())

d_1 <- c(2, 4, 6, 8)
d_2 <- c(5, 15, 2, 12)

m_1 <- matrix(d_1, nrow = 2, ncol = 2)
m_2 <- matrix(d_2, nrow = 2, ncol = 2)

# Element by Element Multiplication

m_1_2 <- m_1 * m_2

kable(m_1_2,
      caption = "Element by Element Multiplication (M1*M2)") %>%
kable_styling(font_size = 12,
              position  = "center",
              full_width = FALSE)

```

## Matrix Multiplication

If **A** is a $(m \times n)$ matrix and **B** is a $(n \times p)$ matrix, then we can form the matrix product **C** by **AB**, if and only if the two matrices are compatible.

The first step is to ensure the matrices are compatible, that is, the inner dimensions of the matrices must match. In this case, the columns in **A** are equal to the rows in **B**, so the matrices are compatible.

The resulting matrix will have the dimensions of the 'outer' numbers. If we examine this from a dimensional perspective then:

$$
A \times B = (m \times n) \times (n \times p) = (m \times p)
$$

For example, assume that **A** is $(5 \times 5)$ and **B** is $(5 \times 8)$. We can form the matrix product **AB** since the as the number of columns in **A** is equal to the number of rows in **B.\
\
*We cannot, however, form the matrix product BA since the number of columns in B is not equal to the number of rows in A****.*

*In general, the only time we are assured that we can multiply AB and BA is when A and B are square matrices with the same dimensions.*

Recall that we have previously defined two matrices as:

$$
m_1 = \begin{bmatrix}
2 & 6 \\
4 & 8 
\end{bmatrix}$$

$$m_2 = \begin{bmatrix}
5 & 2 \\
15 & 12 
\end{bmatrix}
$$

Step 1: **Top-Left element**: Take the first row of $m_1$ and multiply it by the first column of $m_2$.

$$
(2 \times 5) + (6 \times 15) = 10 + 90 = 100
$$

Step 2: **Top-Right Element**: Take the first row of $m_1$ and multiply it by the second column of $m_2$.

$$
(2 \times 2) + (6 \times 12) = 4 + 72 = 76
$$

Step 3: **Bottom-Left element**: Take the second row of $m_1$ and multiply it by the first column of $m_2$.

$$
(4 \times 5) + (8 \times 15) = 20 + 120 = 140
$$

Step 4: **Bottom-right element**: Take the second row of $m_1$ and multiply it by the second column of $m_2$.

$$
(4 \times 2) + (8 \times 12) = 8 + 96 = 104
$$

When we multiply $m_1$ by $m_2$ we obtain a new matrix that is the matrix product or

$$
m_1m_2 = \begin{bmatrix}
2 & 6 \\
4 & 8 
\end{bmatrix}
\begin{bmatrix}
5 & 2 \\
15 & 12 
\end{bmatrix}
=
\begin{bmatrix}
(2 \times 5) + (6 \times 15) & (2 \times 2) + (6 \times 12) \\
(4 \times 5) + (8 \times 15) & (4 \times 2) + (8 \times 12)
\end{bmatrix}
=
\begin{bmatrix}
100 & 76 \\
140 & 104
\end{bmatrix}
$$

To illustrate the point that, in general, matrix multiplication depends on the order of the matrices, we can calculate the product of $m_2 m_1$.

Step 1: **Top-Left element**: Take the first row of $m_2$ and multiply it by the first column of $m_1$.

$$ (5 \times 2) + (2 \times 4) = 10 + 8 = 18 $$

Step 2: **Top-Right Element**: Take the first row of $m_2$ and multiply it by the second column of $m_1$.

$$ (5 \times 6) + (2 \times 8) = 30 + 16 = 46 $$

Step 3: **Bottom-Left element**: Take the second row of $m_2$ and multiply it by the first column of $m_1$.

$$ (15 \times 2) + (12 \times 4) = 30 + 48 = 78 $$

Step 4: **Bottom-right element**: Take the second row of $m_1$ and multiply it by the second column of $m_2$.

$$ (15 \times 6) + (12 \times 8) = 186 $$

When we multiply $m_1$ by $m_2$ we obtain a new matrix that is the matrix product or

$$ m_2 m_1 = \begin{bmatrix} 5 & 2 \\ 15 & 12  \end{bmatrix} \begin{bmatrix} 2 & 6 \\ 4 & 8  \end{bmatrix} = \begin{bmatrix} (5 \times 2) + (2 \times 4) & (5 \times 6) + (2 \times 8) \\ (15 \times 2) + (12 \times 4) & (15 \times 6) + (12 \times 8) \end{bmatrix} = \begin{bmatrix} 18 & 46 \\ 78 & 186 \end{bmatrix} $$

**This illustrates the point: matrix multiplication is non-communicative. Comparability and order of operations matter. Change the order, obtain a different result.**

```{webr-r}

rm(list = ls())

d_1 <- c(2, 4, 6, 8)
d_2 <- c(5, 15, 2, 12)

m_1 <- matrix(d_1, nrow = 2, ncol = 2)
m_2 <- matrix(d_2, nrow = 2, ncol = 2)

# Matrix Multiplication

mm_1_2 <- m_1 %*% m_2

kable(mm_1_2,
      caption = "Matrix Multiplication (M1 %*% M2") 


# Matrix Multiplication

mm_2_1 <- m_2 %*% m_1

kable(mm_2_1,
      caption = "Matrix Multiplication (M2 %*% M1") 

```

## Matrix Inversion

To understand the inverse of a matrix, let's start with an example where have an equation and we want to solve for the unknown value of $x$.

$$
5x = 10
$$

We can divide both sides by the reciprocal of 5 to solve for $x$. $$
\frac{1}{5} \times 5x = 10 \times \frac{1}{5} 
$$In other words, if we multiply 5 by it's inverse, we obtain

$$
5 \times \frac{1}{5} = 1
$$

When we multiply a matrix by its inverse, we obtain the **identify matrix (I)** which is the matrix equivalent of the number 1 where the identify matrix has 1's along the diagonal and 0's off the diagonal or:

$$
A A^{-1} = I = 
\begin{bmatrix}1 & 0 & \ldots & 0 \\ 
0 & 1 & \dots & 0 \\
\vdots & & \ddots & \vdots \\
0 & \ldots & \dots & 1
\end{bmatrix}
$$

Not every matrix, however, has an inverse. A matrix must meet two criteria to have an inverse:

-   The matrix must be square, that is, the number of rows and columns are equal

-   The **determinant** cannot be zero.

Define a $(2 \times 2)$ matrix as:

$$X = \begin{bmatrix}
A & B \\
C & D
\end{bmatrix}$$

The inverse of the $X$ matrix is denoted as $X^{-1}$ and is equal to:

$$X^{-1} = \frac{1}{AD - BC} \begin{bmatrix}
D & -B \\
-C & A
\end{bmatrix}$$

We should note that $(AD - BC)$ is the **determinant**.

Now define $X$ as:

$$X = \begin{bmatrix}
4 & 8 \\
2 & 6
\end{bmatrix}$$

The inverse of $X$ or $X^{-1}$ is equal to:

$$X^{-1} = \frac{1}{4*6 - 8*2} \begin{bmatrix}
6 & -8\\
-2 & 4
\end{bmatrix}$$

$$X^{-1} = \frac{1}{8} \begin{bmatrix}
6 & -8\\
-2 & 4
\end{bmatrix}$$

$$X^{-1} = \begin{bmatrix}
3/4 & -1\\
-1/4 & 1/2
\end{bmatrix}$$

And if we multiply $X$ by $X^{-1}$, we would obtain:

$$X X^{-1} = \begin{bmatrix}
4 & 8 \\
2 & 6
\end{bmatrix} 
\begin{bmatrix}
3/4 & -1\\
-1/4 & 1/2
\end{bmatrix} = 
\begin{bmatrix}
(4*3/4 + -1/4*8) & (4*-1 + 8*1/2) \\ 
(2*3/4 + 6*-1/4) & (2*-1 + 6*1/2)
\end{bmatrix}
=
\begin{bmatrix}
1 & 0 \\
0 & 1
\end{bmatrix}=
I$$

```{r c4, message = FALSE}

# Create Matrix M3

m_3 <- matrix(c(4,2,8,6), nrow = 2, ncol = 2)

kable(m_3,
      caption = "Matrix M3") %>%
kable_styling(font_size = 12,
              position  = "center",
              full_width = FALSE)

# Create the inverse of M3

inv_m3 <- solve(m_3)

kable(inv_m3,
      caption = "Inverse of Matrix M3") %>%
kable_styling(font_size = 12,
              position  = "center",
              full_width = FALSE)


# Matrix Multiplication M3 * inverse M3
# Yields Identity Matrix

m_3_inv_m3 <- m_3 %*% solve(m_3)

inv_m3_m3 <- solve(m_3) %*% m_3

kable(m_3_inv_m3,
      caption = "M3 times Inverse M3") %>%
kable_styling(font_size = 12,
              position  = "center",
              full_width = FALSE)


```
