---
title: "R Squared"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Introduction

In evaluating the performance of Ordinary Least Squares regressions, the
total variation of the dependent variable is measured by the Total Sum
of Squares (SST), which is calculated using the symmetric and idempotent
centering matrix. This framework allows for the decomposition of SST
into the Explained Sum of Squares (SSE) and the Residual Sum of Squares
(SSR). From this decomposition, the coefficient of determination, $R^2$
is derived to represent the proportion of total variation explained by
the independent variables. However, $R^2$ is only bounded between 0 and
1 when the model includes a constant term; without an intercept, the
mathematical properties that simplify the SST decomposition fail to
hold, making an unpredictable and unreliable metric.

A significant limitation of $R^2$ is its mathematical property of
non-decrease; adding any additional explanatory variable to a model will
mechanically increase $R^2$, even if the variable has no meaningful
relationship with the dependent variable. To correct for this, the
Adjusted $R^2$ incorporates a penalty for each added regressor based on
the degrees of freedom. Unlike the standard $R^2$, the adjusted version
will only increase if the new variable’s contribution to the model's
explanatory power is sufficient to offset the loss in degrees of
freedom.

## Total Sum of Squares for x

We can use $\boldsymbol{M^0}$ to calculate the sum of the squared mean
deviations of $\boldsymbol{x}$.

For one variable $\boldsymbol{x}$, the sum of squared mean deviations
is:

$$\sum_{i=1}^{n}(x_i - \bar{x})^2 = (x_i - \bar{x})^T(x_i - \bar{x}) = (M^0 x)^T (M^0 x)$$

$$\sum_{i=1}^{n}(x_i - \bar{x})^2 = (M^0 x)^T (M^0 x) = x^T M^{0T}M^0x$$
Recall that $M^0$ is symmetric or

$$M^{0T} = M^0$$

Recall that $M^0$ is idempotent or

$$M^{0T}M^{0} = M^0$$

The sum of squared mean deviations for a single variable becomes:

$$\sum_{i=1}^{n}(x_i - \bar{x})^2  = x^T M^{0T}M^0x = x^T M^0 x$$

## Total Sum of Squares for X

First, the transformation of the $\boldsymbol{X}$ matrix into mean
deviations is:

$$
\boldsymbol{M^0 X}
$$

The sum of squares and cross-products of the mean deviations for
$\boldsymbol{X}$ is:

$$
\boldsymbol{(M^0 X)^T (M^0 X)}
$$

$$
\boldsymbol{X^T M^{0T} M^0 X}
$$

Since $\boldsymbol{M^0}$ is symmetric and idempotent, this can be
simplified to:

$$
\boldsymbol{X^T M^0 X}
$$

While the single variable case for $\boldsymbol{x^T M^0 x}$ yields a
scalar which is the Total Sum of Squares for that variable, the matrix
operation yields:

$$ \boldsymbol{X^T M^0 X} \sim (K \times K) $$

The diagonal elements of $\boldsymbol{X^T M^0 X}$ are the sum of the
squared deviations (SST) for each individual variable in the
$\boldsymbol{X}$ matrix. The off-diagonal elements represent the sum of
the squared cross-products (SSCP) between different variables and are
used to calculate covariances or:

$$
\boldsymbol{X^T M^0 X} = \begin{bmatrix} SST_{x1} & SSCP_{x1,x2} & \dots & SST_{x1,xK} \\
SSCP_{x2,x1} & SST_{x2} & \dots & SSCP_{x2,xK} \\
\vdots & \ddots & \dots & \vdots \\
SSCP_{xk,x1} & SST_{xk,x2} & \dots & SST_{xk} \\
\end{bmatrix}
$$

However, recall that the first column of $\boldsymbol{X}$ consists of
1's and so every observation of $\boldsymbol{x_1}$ is equal to its mean
and thus the mean deviation of every observation in $\boldsymbol{x_1}$
is 0. Also, since the off-diagonal elements represent the shared
variation between variables (cross-products) and since
$\boldsymbol{x_1}$ does not vary, the sum of the squared cross-product
between $\boldsymbol{x_1}$ and any other variable is 0.

$$
\boldsymbol{X^T M^0 X} = \begin{bmatrix} 0 & 0 & \dots & 0 \\
0 & SST_{x2} & \dots & SSCP_{x2,xK} \\
\vdots & \ddots & \dots & \vdots \\
0 & SST_{xk,x2} & \dots & SST_{xk} \\
\end{bmatrix}
$$

## Sum of Squares

For the dependent variable $y$, the **total sum of squares**, $SST_y$,
is equal to the sum of squared mean deviations.

$$
SST_y = \boldsymbol{y^T M^0 y}
$$

Given the relationship, $\boldsymbol{y = Xb + e}$, we can pre-multiply
by the centering matrix, $\boldsymbol{M^0}$ or:

$$
\boldsymbol{y = Xb +e}
$$

$$
\boldsymbol{M^0 y = M^0 Xb + M^0e}
$$

If the model contains a constant term and the other assumptions hold,
then the mean of the residuals is equal to zero, that is,
$\boldsymbol{\bar{e} = 0}$ this implies that:

$$
\boldsymbol{M^0 e = e}
$$\
$$
\boldsymbol{M^0 y = M^0 Xb + e}
$$

And we can pre-multiply by $\boldsymbol{y^T}$ to obtain:

$$
\boldsymbol{y^T M^0 y = (M^0 Xb +e)^T  (M^{0}Xb + e)}
$$

$$
\boldsymbol{y^T M^0 y = b^T X^T M^{0T}M^0Xb + b^T X^T M^{0T}e + e^TM^0Xb + e^Te}
$$

Because the residuals, $\boldsymbol{e}$ are orthogonal to the
regressors, $\boldsymbol{X}$ and, by extension, to the centered
regressors, $\boldsymbol{M^0X}$, the cross-product terms drop out. We
can also rely on the fact that the centering matrix is symmetric and
idempotent.

This leaves us with a fundamental identity:

$$
\boldsymbol{y^T M^0 y = b^T X^T M^0 Xb + e^T e}
$$

Which can be expressed as:

$$
SST = SSE + SSR
$$

That is, the total sum of squares of the dependent variable (SST) is
equal to the explained sum of squares (SSE) and the residual sum of
squares (SSR).

## Decomposition in R

Building upon our previous examples, we can use simulated data to create
our independent variables, population coefficients, and the dependent
variable. We can estimate the least squares coefficients and construct
the projection and residual maker matrices.

We then construct the $\boldsymbol{M^0}$ matrix and use this matrix to
estimate SST, SSE, and SSR.

The output table demonstrates the equality, that the total sum of
squares can be decomposed into the explained sum of squares and the
residual sum of squares.

```{r}

rm(list = ls())
library(kableExtra)

nrows <- 2000
ncols <- 5
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- rnorm(nrows)
        gc()
      }

beta <- seq(1,ncols)

y <- x %*% beta + rnorm(nrows)

bhat <- solve(t(x) %*% x) %*% t(x) %*% y

I <- diag(nrows)
P <- x %*% solve(t(x) %*% x) %*% t(x)
M <- I - x %*% solve(t(x) %*% x) %*% t(x)

i <- matrix(1,nrows,1)

M0 <- I - 1/nrows * i %*% t(i)

yhat <- P %*% y
e <- M %*% y


sst <- t(y) %*% M0 %*% y
sse <- t(bhat) %*% t(x) %*% M0 %*% x %*% bhat
ssr <- t(e) %*% e

df <- tibble(SST = as.numeric(sst),
             SSE = as.numeric(sse),
             SSR = as.numeric(ssr))

kable(df,
      col.names = c('SST', 'SSE', 'SSR'),
      align     = 'c',
      digits    = 2) %>%
kable_classic()

```

## R-Squared

The total sum of squares (SST) can be decomposed into the regression (or
explained) sum of squares (SSE) and the residual sum of squares (SSR).

$$\boldsymbol{y^T M^0 y} = (\boldsymbol{b^T X^T M^0Xb}) + (\boldsymbol{e^T e})$$
$$SST = SSE + SSR$$The **coefficient of determination** is a measure of
how well a regression fits the data, where:
$$R^2 = \frac{SSE}{SST} = \frac{(\boldsymbol{b^T X^T M^0Xb})}{\boldsymbol{y^T M^0 y}} = 1 - \frac{\boldsymbol{e^Te}}{\boldsymbol{y^T M^0 y}} = 1 - \frac{SSR}{SST}$$

$$
0 \le R^2 \le 1
$$

When $R^2 = 0$ the predicted values of $y$ are $\bar{y}$ so
$\boldsymbol{X}$ has no explanatory power.

When $R^2 = 1$, the predicted values and actual values are the same.

## R-Squared in R

First, we can show that $\boldsymbol{M^0 e = e}$.

We can then calculate the total sum of squares, estimated sum of
squares, and residual sum of squares.

$R^2$ is equivalent whether obtained using the estimated or residual sum
of squares.

One problem, however, is that one can continue to increase $R^2$ by
adding explanatory variables to the regression model.

```{r}

rm(list = ls())
library(kableExtra)

nrows <- 2000
ncols <- 5
 
x <- matrix(0, nrows, ncols)
      x[,1] <- 1
      for (l in 2: ncols) {
        x[,l] <- rnorm(nrows)
        gc()
      }

beta <- seq(1,ncols)

y <- x %*% beta + rnorm(nrows)

bhat <- solve(t(x) %*% x) %*% t(x) %*% y

I <- diag(nrows)
P <- x %*% solve(t(x) %*% x) %*% t(x)
M <- I - x %*% solve(t(x) %*% x) %*% t(x)

i <- matrix(1,nrows,1)

M0 <- I - 1/nrows * i %*% t(i)

yhat <- P %*% y
e <- M %*% y

y_cal <- yhat + e

M0_e <- M0 %*% e

e[1:4]
M0_e[1:4]

SST <- t(y) %*% M0 %*% y
SSE <- t(bhat) %*% t(x) %*% M0 %*% x %*% bhat
R2 <- SSE / SST

R2_r = 1 - (t(e) %*% e) / (t(y) %*% M0 %*% y)

cbind(R2, R2_r)

```

## R-Squared Increases in X

When we add a new variable $\boldsymbol{z}$ to a regression already
containing $\boldsymbol{X}$, we can express the new residual sum of
squares in terms of the old residual sum of squares or:

$$
u^T u = e^T e (1- r_{yz}^2)
$$ where $r_{yz}^2$ is the squared correlation between $\boldsymbol{y}$
and $\boldsymbol{z}$ after both have been 'purged' of the influence of
$\boldsymbol{X}$.

Since $r^2$ must be between 0 and 1, it follows that

$$
u^T u \le e^T e
$$ In terms of $R^2$, this relationship simplifies to:

$$
R_{Xz}^2 = R_X^2 +(1-R_X^2)r_{yz}^2
$$ What does this mean?

The new $R^2$ will always be greater than or equal to the old $R^2$.

**In other words, adding an additional explanatory variable will never
lower** $R^2$**.**

## Constant Term Required

For $R^2$ to be a reliable measure bounded between 0 and 1, the
$\boldsymbol{X}$ matrix must include a constant term (the column of
1's).

With a constant term, we know:

$$
\boldsymbol{M^0 e = e}
$$ $$
\boldsymbol{e^T X = 0}
$$Without a constant term, the mean of the residuals is not zero and the
residuals are no longer orthogonal to the centered regressors.

Without an intercept, the cross-product term does not drop out. This
breaks the SST=SSE+SSR identity, potentially leading to an
"unpredictable" R2 that can even be negative.

## Adjusted R-Squared

Given that one can increase $R^2$ to 1 by adding independent variables,
we can use **adjusted** $R^2$ to examine the impact of the addition (or
deletion) of an independent variable.

$$\bar{R^2} = 1 - \frac{\boldsymbol{e^Te}}{\boldsymbol{y^T M^0 y}}\frac{(n-1)}{(n-K)} = 1 - (1-R^2)\frac{(n-1)}{(n-K)} $$
$\bar{R^2}$ will fall (rise) when the variable $x$ is deleted from the
regression if the square of the t-ratio associated with this variable is
greater (less) than 1.

In other words, the greater the statistical significance of a variable
that is removed, the more significant the fall in adjusted $R^2$.

The standard statistical output will include both $R^2$ and $\bar{R^2}$
in the output.

## Example

In the results to the right, we regress reading scores by school on the
school characteristics.

We then add a variable that is statistically significant.

Note how adding this variable increases $R^2$ and $\bar{R^2}$.

We then add a variable that is not statistically significant.

While $R^2$ increases, $\bar{R^2}$ does not.

```{r, message = FALSE, warings = FALSE}

rm(list = ls())
library(AER)
library(modelsummary)
data(CASchools)

mod_2 <- lm(read ~ income + lunch + teachers, data = CASchools)
mod_3 <- lm(read ~ income + lunch + english, data = CASchools)
mod_4 <- lm(read ~ income + lunch + english + computer, data = CASchools)

modelsummary(list(mod_2, mod_3, mod_4))
```
